---
title: "NCOS Artificial Refuges Study - Revisions"
author: "Garrett Craig"
date: "`r Sys.Date()`"
format:
  html:
    toc: true
    toc-location: left
    toc-depth: 3
    number_sections: true
    code-fold: true
    code-tools: true
    code-summary: "Show Code"
    embed-resources: true
    theme: darkly
    page-layout: full
    html-math-method: katex
    fig-width: 10
    fig-height: 6
    fig-format: png
    fig-responsive: true
  pdf:
    toc: true
    number-sections: true
    fig-width: 10
    fig-height: 6
    fig-format: png
    pdf-engine: lualatex
    mainfont: "Palatino"
    code-block: false   # ← hides code blocks from the PDF
    execute:
      echo: false
execute:
  eval: true
  message: false
  warning: false
editor:
  markdown:
    wrap: sentence
---

# Setup
```{r include=FALSE}
#clean your environment
rm(list=ls())

# Render NA as an empty cell in every kable() table rather than the literal
# string "NA" (e.g. ratio columns that are intentionally blank for intercept rows).
options(knitr.kable.NA = "")

# Core utility libraries
library(here)          # file path management
library(janitor)       # clean variable names
library(tidyverse)     # includes ggplot2, dplyr, readr, purrr, tibble, etc.
library(lubridate)     # date-time parsing (not included in tidyverse)
library(broom)         # tidy model outputs
library(knitr)         # reporting and markdown
library(readxl)        # reading Excel files
library(kableExtra)
library(stargazer)

# Data visualization enhancements
library(ggtext)        # rich text in ggplot
library(ggsignif)      # significance bars for ggplot
library(patchwork)     # ggplot combining
library(scales)        # axis scaling and labeling
library(webshot2)

# Statistical analysis
library(effsize)       # for Cohen’s d and other effect sizes
library(simpleboot)    # for bootstrapping
library(boot)          # bootstrapping (more general)
library(car)           # regression tools, including ANOVA, VIF
library(MASS)          # negative binomial models
library(MuMIn)         # AICc and model selection
library(spdep)         # spatial autocorrelation (Moran’s I)
library(dplyr)        # reloaded to ensure dplyr’s functions are used

# Mapping and spatial
library(leaflet)       # interactive maps
library(RColorBrewer)  # color palettes for plots and maps
library(maptiles)
library(terra)
library(sf)
library(ggplot2)
library(cowplot)
library(maps)
library(ggspatial)
library(ggnewscale)    # for multiple fill scales in ggplot

# Text and string manipulation
library(stringi)       # extended string processing

# Table joins
library(fuzzyjoin)     # fuzzy matching for joins
if (!require("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

BiocManager::install("IRanges")
```

# Part 2: Spring Study

## Load Data
```{r}
##load study part 2 data from April/May 2021
raw_pt_2_seq = read_csv(here("data","sequences.csv"))
raw_pt_2_dep = read_csv(here("data","deployments.csv"))

# Read in the site-to-habitat crosswalk
site_to_habitat_crosswalk <- read_csv(here("data", "site_to_habitat_crosswalk.csv")) |> 
  mutate(
    trail = factor(trail, levels = c("no", "yes")),
    # Grassland is the reference level: it is the habitat with the highest
    # observed activity, so marsh and scrub read as contrasts against it.
    habitat_type = factor(habitat_type, levels = c("Grassland", "Marsh", "Scrub"))
  )
```
## Clean Data
```{r}
raw_pt_2_dep$feature_type_methodology[raw_pt_2_dep$deployment_id == "L30C9"] <- "Log"
raw_pt_2_dep$feature_type_methodology[raw_pt_2_dep$deployment_id == "H8C11"] <- "Constructed Hibernacula"

# Recode raw CSV value to new terminology
raw_pt_2_dep$feature_type_methodology[raw_pt_2_dep$feature_type_methodology == "Constructed Hibernacula"] <- "Artificial Refuge"

clean_pt_2_seq <- raw_pt_2_seq |> 
  dplyr::select("project_id", "deployment_id", "sequence_id", "is_blank", "identified_by", "wi_taxon_id", 
         "class", "order", "family", "genus", "species", "common_name", "start_time", "end_time", 
         "group_size", "individual_animal_notes", "license") |> 
  clean_names() |> 
  mutate(
    start_time = ymd_hms(start_time),  
    end_time = ymd_hms(end_time),  
    sequence_duration_sec = as.numeric(difftime(end_time, start_time, units = "secs"))  # Sequence duration in seconds
  )

clean_pt_2_dep <- raw_pt_2_dep |>
  dplyr::select(
    deployment_id, placename, longitude, latitude, start_date, end_date,
    feature_type_methodology, camera_id, camera_name, camera_functioning,
    sensor_height, sensor_orientation, remarks
  ) |>
  clean_names() |>
  mutate(
    end_date = case_when(
      deployment_id == "H46C2" ~ as.Date("2021-05-02"),
      deployment_id == "B2C9"  ~ as.Date("2021-04-26"),
      deployment_id == "H7C12" ~ as.Date("2021-05-07"),
      deployment_id == "H35C6" ~ as.Date("2021-05-15"),
      TRUE ~ as.Date(end_date)
    ),
    feature_type_methodology_recoded = factor(
      case_when(
        feature_type_methodology == "Log" ~ "Log",
        feature_type_methodology == "Boulder" ~ "Boulder",
        TRUE ~ "Artificial Refuge"
      ),
      levels = c("Artificial Refuge", "Boulder", "Log")
    ),
    start_date = as.Date(start_date),
    deployment_duration = as.numeric(difftime(end_date, start_date, units = "days")),
    study_part = 2
  ) |>
  filter(deployment_id != "L56") |>
  left_join(site_to_habitat_crosswalk, by = c("placename" = "site"))

# join sequences and deployments data
clean_pt_2_seq_dep <- clean_pt_2_seq |> 
  full_join(clean_pt_2_dep, by = "deployment_id") |> 
  filter(
    class != "No CV Result",
    is.na(genus) | genus != "Homo",    # Remove humans
    class != "Insecta",                # Remove insects
    species != "catus"                 # Remove domestic cats
  ) |> 
  mutate(
    genus_species = paste(genus, species, sep = " "),
    start_time = ymd_hms(start_time),
    start_date = as.Date(start_date),
    end_date = as.Date(end_date),
    feature_type_methodology = as.factor(feature_type_methodology),
    obs_start_date = as.Date(start_time),
    obs_start_time = hms::as_hms(start_time),
    feature_type_methodology_recoded = case_when(
      feature_type_methodology == "Log" ~ "Log",
      feature_type_methodology == "Boulder" ~ "Boulder",
      TRUE ~ "Artificial Refuge"
    ),
    feature_type_methodology_recoded = factor(
      feature_type_methodology_recoded,
      levels = c("Artificial Refuge", "Boulder", "Log")
    )
  )
```

```{r}
### Remove Duplicate Sequences

# Step 1: Add row_id based on existing sequence_id
data_intervals_pt_2 <- clean_pt_2_seq_dep %>%
  mutate(row_id = sequence_id)

# Step 2: Self-join to find overlapping intervals (same species, place, date; different cameras)
overlap_pairs_pt_2 <- interval_inner_join(
 data_intervals_pt_2, data_intervals_pt_2,
  by = c("start_time", "end_time"),
  maxgap = 0,
  type = "any"
) %>%
  filter(
    wi_taxon_id.x == wi_taxon_id.y,
    placename.x == placename.y,
    obs_start_date.x == obs_start_date.y,
    camera_name.x != camera_name.y,
    row_id.x != row_id.y
  ) %>%
   dplyr::select(
    wi_taxon_id.x, wi_taxon_id.y,
    placename.x, placename.y,
    obs_start_date.x, obs_start_date.y,
    camera_name.x, camera_name.y,
    row_id.x, row_id.y,
    common_name.x, common_name.y,
    group_size.x, group_size.y,
    start_time.x, start_time.y,
    end_time.x, end_time.y
  )

# Step 3: Decide which rows to remove (keep higher group_size; if tied, keep later end_time)
rows_to_remove_pt_2 <- overlap_pairs_pt_2 %>%
  mutate(
    keep_x = if_else(group_size.x > group_size.y, TRUE,
              if_else(group_size.x < group_size.y, FALSE,
                end_time.x > end_time.y))  # If tie, later end_time wins
  ) %>%
  filter(!keep_x) %>%
  pull(row_id.x) %>%
  unique()

# Step 4: Remove the lower-priority overlaps
clean_pt_2_seq_dep <- clean_pt_2_seq_dep %>%
  filter(!sequence_id %in% rows_to_remove_pt_2)
```

## Summarize

### Calculate trapping effort by placename
```{r}
#effort
placename_effort_summary_pt_2 <- clean_pt_2_dep %>%
  group_by(placename,longitude,latitude,feature_type_methodology, feature_type_methodology_recoded,study_part,habitat_type,trail,notable_entrances) %>%
  summarise(
    total_camera_days = sum(deployment_duration, na.rm = TRUE),
    total_camera_hours = total_camera_days * 24,
    .groups = "drop"
  )

#total detections
placename_detections_summary_pt_2 <- clean_pt_2_seq_dep %>%
  group_by(placename) %>%
  summarise(
    total_detections = sum(group_size, na.rm = TRUE),
    .groups = "drop"
  )

#join effort and detections
placename_summary_pt_2 <- placename_effort_summary_pt_2 %>%
  left_join(placename_detections_summary_pt_2, by = "placename") %>%
  mutate(
    detections_per_camera_day = total_detections / total_camera_days,
    detections_per_camera_hour = total_detections / total_camera_hours
  ) %>%
  arrange(desc(total_detections))
```


## Statistical Analysis (Revised)

### Negative Binomial GLMs
```{r}
#| echo: false
# ============================================================================
# REVISED ANALYSIS addressing reviewer feedback:
#   1) Spatial autocorrelation testing (Moran's I)
#   2) Interaction effects
#   3) AICc instead of AIC (small-sample correction, n=27)
#   4) Prefer model with lowest AICc and highest Akaike weight; note parsimonious alternatives
# ============================================================================

# --- Additive models (original set) ---

nb_offset_1a <- MASS::glm.nb(
  total_detections ~ feature_type_methodology_recoded + offset(log(total_camera_hours)),
  data = placename_summary_pt_2)

nb_offset_1b <- MASS::glm.nb(
  total_detections ~ trail + offset(log(total_camera_hours)),
  data = placename_summary_pt_2)

nb_offset_1c <- MASS::glm.nb(
  total_detections ~ habitat_type + offset(log(total_camera_hours)),
  data = placename_summary_pt_2)

nb_offset_2a <- MASS::glm.nb(
  total_detections ~ feature_type_methodology_recoded + trail + offset(log(total_camera_hours)),
  data = placename_summary_pt_2)

nb_offset_2b <- MASS::glm.nb(
  total_detections ~ feature_type_methodology_recoded + habitat_type + offset(log(total_camera_hours)),
  data = placename_summary_pt_2)

nb_offset_2c <- MASS::glm.nb(
  total_detections ~ trail + habitat_type + offset(log(total_camera_hours)),
  data = placename_summary_pt_2)

nb_offset_3a <- MASS::glm.nb(
  total_detections ~ feature_type_methodology_recoded + trail + habitat_type + offset(log(total_camera_hours)),
  data = placename_summary_pt_2)

# --- Interaction models (NEW - reviewer request) ---

nb_offset_int_ft_hab <- tryCatch(
  MASS::glm.nb(
    total_detections ~ feature_type_methodology_recoded * habitat_type + offset(log(total_camera_hours)),
    data = placename_summary_pt_2),
  error = function(e) { message("Feature Type * Habitat interaction model failed to converge: ", e$message); NULL }
)

nb_offset_int_ft_trail <- tryCatch(
  MASS::glm.nb(
    total_detections ~ feature_type_methodology_recoded * trail + offset(log(total_camera_hours)),
    data = placename_summary_pt_2),
  error = function(e) { message("Feature Type * Trail interaction model failed to converge: ", e$message); NULL }
)

nb_offset_int_ft_trail_hab <- tryCatch(
  MASS::glm.nb(
    total_detections ~ feature_type_methodology_recoded * trail + habitat_type + offset(log(total_camera_hours)),
    data = placename_summary_pt_2),
  error = function(e) { message("Feature Type * Trail + Habitat interaction model failed to converge: ", e$message); NULL }
)

# --- Estimability check -------------------------------------------------------
# An interaction is only estimable when every combination of factor levels is
# observed. Here all five trail-adjacent sites are artificial refuges (there are
# no trail-adjacent boulders or logs), so the Boulder:trail and Log:trail cells
# are empty. glm.nb still "converges" but silently returns NA for those aliased
# coefficients, and the model collapses to its additive counterpart. Left in the
# candidate set it would duplicate an existing model and split its Akaike weight.
is_estimable <- function(model) {
  !is.null(model) && !any(is.na(stats::coef(model)))
}

# Cross-tabulate the interacting factors to document the empty cells
cat("Feature type x trail crosstab (sites):\n")
print(with(placename_summary_pt_2, table(feature_type_methodology_recoded, trail)))

cat("\nInteraction model status:\n")
for (nm in c("Feature Type x Habitat", "Feature Type x Trail", "Feature Type x Trail + Habitat")) {
  mod <- switch(nm,
    "Feature Type x Habitat"         = nb_offset_int_ft_hab,
    "Feature Type x Trail"           = nb_offset_int_ft_trail,
    "Feature Type x Trail + Habitat" = nb_offset_int_ft_trail_hab)

  if (is.null(mod)) {
    cat(sprintf("  %-32s FAILED TO CONVERGE\n", nm))
  } else if (!is_estimable(mod)) {
    aliased <- names(coef(mod))[is.na(coef(mod))]
    cat(sprintf("  %-32s RANK DEFICIENT - dropped (aliased: %s)\n",
                nm, paste(aliased, collapse = ", ")))
  } else {
    cat(sprintf("  %-32s estimable\n", nm))
  }
}
```

### AICc Model Comparison (small-sample corrected)
```{r}
#| echo: false
# Build named list of all successfully fitted models
all_models <- list(
  "Feature Type" = nb_offset_1a,
  "Trail" = nb_offset_1b,
  "Habitat Type" = nb_offset_1c,
  "Feature Type + Trail" = nb_offset_2a,
  "Feature Type + Habitat Type" = nb_offset_2b,
  "Trail + Habitat Type" = nb_offset_2c,
  "Feature Type + Trail + Habitat Type" = nb_offset_3a
)

# Add interaction models only if they converged AND are fully estimable.
# Rank-deficient interactions are excluded: their aliased terms are dropped by
# glm.nb, leaving a model identical to its additive counterpart, which would
# otherwise enter the candidate set twice and split the Akaike weight.
if (is_estimable(nb_offset_int_ft_hab))
  all_models[["Feature Type x Habitat Type"]] <- nb_offset_int_ft_hab
if (is_estimable(nb_offset_int_ft_trail))
  all_models[["Feature Type x Trail"]] <- nb_offset_int_ft_trail
if (is_estimable(nb_offset_int_ft_trail_hab))
  all_models[["Feature Type x Trail + Habitat"]] <- nb_offset_int_ft_trail_hab

# Compute AICc for each model (small-sample correction, n = 27)
aicc_table <- data.frame(
  Model = names(all_models),
  df = sapply(all_models, function(m) attr(logLik(m), "df")),
  AICc = sapply(all_models, MuMIn::AICc),
  stringsAsFactors = FALSE
)
aicc_table$delta_AICc <- aicc_table$AICc - min(aicc_table$AICc)

# Akaike weights
aicc_table$rel_likelihood <- exp(-0.5 * aicc_table$delta_AICc)
aicc_table$akaike_weight <- aicc_table$rel_likelihood / sum(aicc_table$rel_likelihood)

aicc_table <- aicc_table %>%
  arrange(delta_AICc) %>%
  dplyr::select(Model, df, AICc, delta_AICc, akaike_weight)

cat("=== AICc Model Comparison ===\n")
print(aicc_table, digits = 3, row.names = FALSE)

# Sort models to match AICc ranking
model_order <- match(aicc_table$Model, names(all_models))
sorted_models <- all_models[model_order]
sorted_delta_aicc <- aicc_table$delta_AICc
sorted_weights <- aicc_table$akaike_weight

# Identify best model: lowest AICc (= highest Akaike weight)
competitive_idx <- which(sorted_delta_aicc < 2)
best_idx <- competitive_idx[1]
best_model <- sorted_models[[best_idx]]
best_model_name <- names(sorted_models)[best_idx]

# Most parsimonious alternative within ΔAICc < 2
competitive_df <- sapply(sorted_models[competitive_idx], function(m) attr(logLik(m), "df"))
parsim_idx <- competitive_idx[which.min(competitive_df)]
parsim_name <- names(sorted_models)[parsim_idx]

cat("\n=== Model Selection ===\n")
cat("Models within ΔAICc < 2:\n")
for (i in competitive_idx) {
  cat(sprintf("  %s (df=%d, ΔAICc=%.2f, w=%.3f)\n",
    names(sorted_models)[i],
    attr(logLik(sorted_models[[i]]), "df"),
    sorted_delta_aicc[i],
    sorted_weights[i]))
}

cat(sprintf("\nPreferred model (lowest AICc, highest Akaike weight): %s (w = %.3f)\n",
    best_model_name, sorted_weights[best_idx]))

if (parsim_name != best_model_name) {
  cat(sprintf("Most parsimonious alternative: %s (df=%d, w=%.3f)\n",
      parsim_name, attr(logLik(sorted_models[[parsim_idx]]), "df"),
      sorted_weights[parsim_idx]))
}

er <- sorted_weights[best_idx] / sorted_weights[competitive_idx[2]]
cat(sprintf("Evidence ratio (preferred / next competitive): %.2f\n", er))

cat("\n=== Model-Averaged Coefficients (MuMIn::model.avg) ===\n")
cat("Weighted by Akaike weights across competitive set (ΔAICc < 2).\n")
cat("Note: model averaging is common but debated in ecology (see Cade 2015, Banner & Higgs 2017).\n\n")

top_idx <- competitive_idx
top_models <- sorted_models[top_idx]

model_avg_result <- tryCatch({
  mod_sel <- model.sel(top_models, rank = "AICc")
  model.avg(mod_sel, revised.var = TRUE)
}, error = function(e) {
  message("MuMIn::model.avg failed: ", e$message)
  NULL
})

if (!is.null(model_avg_result)) {
  cat("--- Conditional averages (averaged only over models containing each term) ---\n")
  print(summary(model_avg_result))

  cond_coefs <- as.data.frame(coefTable(model_avg_result, full = FALSE))
  cond_coefs$Coefficient <- rownames(cond_coefs)
  rownames(cond_coefs) <- NULL

  avg_display <- cond_coefs %>%
    # Terms that could not be averaged (e.g. interaction terms present in a single
    # model) come back as NA; drop them rather than printing empty rows.
    filter(!is.na(Estimate)) %>%
    mutate(
      Predictor = case_when(
        Coefficient == "(Intercept)" ~ "Intercept (Artificial Refuge, Grassland, Interior)",
        Coefficient == "feature_type_methodology_recodedBoulder" ~ "Boulder",
        Coefficient == "feature_type_methodology_recodedLog" ~ "Log",
        Coefficient == "trailyes" ~ "Trail-adjacent",
        Coefficient == "habitat_typeMarsh" ~ "Marsh",
        Coefficient == "habitat_typeScrub" ~ "Scrub",
        Coefficient == "feature_type_methodology_recodedBoulder:trailyes" ~ "Boulder × Trail-adjacent",
        Coefficient == "feature_type_methodology_recodedLog:trailyes" ~ "Log × Trail-adjacent",
        TRUE ~ Coefficient
      ),
      # As in Table 1: exp(beta) is a rate ratio only for contrast terms. The
      # intercept exponentiates to the baseline rate itself, so the ratio, %
      # change and ratio CI columns are left blank for that row.
      is_intercept = Coefficient == "(Intercept)",
      IRR = ifelse(is_intercept, NA_real_, exp(Estimate)),
      Pct_Change = ifelse(is_intercept, NA_real_, round((exp(Estimate) - 1) * 100, 1)),
      CI_Lower = ifelse(is_intercept, NA_real_, exp(Estimate - 1.96 * `Std. Error`)),
      CI_Upper = ifelse(is_intercept, NA_real_, exp(Estimate + 1.96 * `Std. Error`))
    )

  # Baseline rate for the reference condition, reported in the caption.
  baseline_rate_hr  <- exp(avg_display$Estimate[avg_display$is_intercept])
  baseline_rate_day <- baseline_rate_hr * 24

  avg_display %>%
    dplyr::select(Predictor, Estimate, `Std. Error`, IRR, Pct_Change, CI_Lower, CI_Upper) %>%
    kable(
      col.names = c("Predictor", "β", "SE", "IRR", "% Change", "IRR 95% CI Lower", "IRR 95% CI Upper"),
      digits = 3,
      caption = sprintf("Model-Averaged Coefficients (conditional averaging via MuMIn, %d competitive models, ΔAICc < 2). Unconditional SEs account for model selection uncertainty. Reference: Artificial Refuge, not trail-adjacent, grassland habitat. Ratio columns are blank for the intercept: exp(β) there is the baseline detection rate (%.3f per camera hour, %.1f per camera day) rather than a ratio.", length(top_models), baseline_rate_hr, baseline_rate_day),
      align = c("l", "r", "r", "r", "r", "r", "r")
    ) %>%
    kable_styling(
      full_width = FALSE,
      bootstrap_options = c("striped", "hover", "condensed"))
}

cat("\nBest model summary:\n")
summary(best_model)
```

### Spatial Autocorrelation Test (Moran's I)
```{r}
#| echo: false
# Addresses reviewer concern: "The authors did not investigate the role of space"

# Create spatial points from site coordinates
coords <- placename_summary_pt_2 %>%
  dplyr::select(longitude, latitude) %>%
  as.matrix()

# Build spatial weights: inverse distance, k=5 nearest neighbors
knn <- spdep::knearneigh(coords, k = 5)
nb_spatial <- spdep::knn2nb(knn)
listw <- spdep::nb2listw(nb_spatial, style = "W")

# Moran's I on all competitive models (ΔAICc < 2)
cat("=== Moran's I Test for Spatial Autocorrelation (all candidate models) ===\n\n")

# Assessed before selection, so run over the full candidate set (see DHARMa note).
competitive_models <- sorted_models
moran_results <- data.frame(
  Model = character(),
  Morans_I = numeric(),
  Expected_I = numeric(),
  P_Value = numeric(),
  stringsAsFactors = FALSE
)

for (i in seq_along(competitive_models)) {
  mod_name <- names(competitive_models)[i]
  mod <- competitive_models[[i]]
  dev_resids <- residuals(mod, type = "deviance")
  moran_test <- spdep::moran.test(dev_resids, listw)

  moran_results <- rbind(moran_results, data.frame(
    Model = mod_name,
    Morans_I = round(moran_test$estimate["Moran I statistic"], 4),
    Expected_I = round(moran_test$estimate["Expectation"], 4),
    P_Value = round(moran_test$p.value, 4),
    stringsAsFactors = FALSE
  ))

  cat(sprintf("--- %s (ΔAICc = %.2f)%s ---\n", mod_name, sorted_delta_aicc[i],
              ifelse(sorted_delta_aicc[i] < 2, "  [competitive]", "")))
  cat(sprintf("  Moran's I = %.4f, p = %.4f %s\n\n",
      moran_test$estimate["Moran I statistic"],
      moran_test$p.value,
      ifelse(moran_test$p.value < 0.05, "** SIGNIFICANT **", "(not significant)")))
}

if (all(moran_results$P_Value >= 0.05)) {
  cat("No significant spatial autocorrelation detected in any candidate model.\n")
  cat("Spatial independence assumption of the GLMs is supported.\n")
} else {
  sig_models <- moran_results$Model[moran_results$P_Value < 0.05]
  cat(sprintf("Significant spatial autocorrelation detected in: %s\n", paste(sig_models, collapse = ", ")))
  cat("Consider adding spatial covariates for these models.\n")
}
```

### Spatial Sensitivity: Edge vs. Interior Sites
```{r}
#| echo: false
# Reviewer flagged that 5 of 14 artificial refuges are on the periphery near trails
# The trail variable already captures this, but we can visualize the spatial pattern

placename_summary_pt_2 <- placename_summary_pt_2 %>%
  mutate(
    site_position = case_when(
      trail == "yes" ~ "Edge (trail-adjacent)",
      TRUE ~ "Interior"
    )
  )

cat("=== Site Position Summary ===\n")
placename_summary_pt_2 %>%
  group_by(feature_type_methodology_recoded, site_position) %>%
  summarise(
    n = n(),
    mean_detections = mean(total_detections, na.rm = TRUE),
    sd_detections = sd(total_detections, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  print()

# Spatial plot of residuals to visually inspect clustering
placename_summary_pt_2$dev_resid_best <- residuals(best_model, type = "deviance")

ggplot(placename_summary_pt_2, aes(x = longitude, y = latitude)) +
  geom_point(aes(color = dev_resid_best, shape = site_position), size = 4) +
  scale_color_gradient2(low = "blue", mid = "white", high = "red", midpoint = 0,
                        name = "Deviance\nResidual") +
  scale_shape_manual(values = c("Edge (trail-adjacent)" = 17, "Interior" = 16),
                     name = "Site Position") +
  labs(
    title = sprintf("Spatial Distribution of Model Residuals (%s)", best_model_name),
    x = "Longitude", y = "Latitude"
  ) +
  theme_cowplot() +
  theme(legend.position = "right")

ggsave(here("figures", "spatial_residuals_map.png"), width = 8, height = 6, dpi = 300, bg = "white")
```

## Tables

```{r}
#| echo: false
# Reusable function to build the detailed model table
create_detailed_model_table <- function(model_list, delta_values, weight_values, observed = NULL) {
  detailed_data <- data.frame()

  for (i in seq_along(model_list)) {
    model <- model_list[[i]]
    model_name <- names(model_list)[i]
    coef_summary <- summary(model)$coefficients
    obs <- if (!is.null(observed)) observed else placename_summary_pt_2$total_detections
    fitted_vals <- fitted(model)
    rmse <- sqrt(mean((obs - fitted_vals)^2))

    detailed_data <- rbind(detailed_data, data.frame(
      Model_Name = model_name,
      Predictor = model_name,
      Beta = NA_real_,
      IRR = NA_real_,
      Pct_Change = NA_real_,
      P_Value = NA_real_,
      Delta_AICc = round(delta_values[i], 2),
      Weight = round(weight_values[i], 3),
      RMSE = round(rmse, 3),
      row_type = "header",
      stringsAsFactors = FALSE
    ))

    for (j in 1:nrow(coef_summary)) {
      predictor_name <- rownames(coef_summary)[j]
      p_value <- coef_summary[j, "Pr(>|z|)"]
      beta <- coef_summary[j, "Estimate"]
      is_intercept <- predictor_name == "(Intercept)"

      # exp(beta) is an incidence rate RATIO only for contrast terms. For the
      # intercept it is the baseline rate itself (detections per camera hour,
      # given the log(camera hours) offset) - there is nothing it is a ratio to,
      # so the ratio and "% change" columns are left blank for that row.
      irr <- if (is_intercept) NA_real_ else exp(beta)
      pct_change <- if (is_intercept) NA_real_ else (exp(beta) - 1) * 100

      stars <- case_when(
        p_value < 0.001 ~ "***",
        p_value < 0.01 ~ "**",
        p_value < 0.05 ~ "*",
        TRUE ~ ""
      )

      clean_predictor <- case_when(
        predictor_name == "(Intercept)" ~ "Intercept (Reference)",
        predictor_name == "feature_type_methodology_recodedBoulder" ~ "Boulder",
        predictor_name == "feature_type_methodology_recodedLog" ~ "Log",
        predictor_name == "trailyes" ~ "Adjacent to Trail",
        predictor_name == "habitat_typeMarsh" ~ "Marsh Habitat",
        predictor_name == "habitat_typeScrub" ~ "Scrub Habitat",
        grepl(":", predictor_name) ~ gsub(
          "feature_type_methodology_recoded|habitat_type|trail", "",
          gsub(":", " x ", predictor_name)),
        TRUE ~ predictor_name
      )

      detailed_data <- rbind(detailed_data, data.frame(
        Model_Name = model_name,
        Predictor = paste0(clean_predictor, stars),
        Beta = round(beta, 3),
        IRR = round(irr, 3),
        Pct_Change = round(pct_change, 1),
        P_Value = round(p_value, 3),
        Delta_AICc = NA_real_,
        Weight = NA_real_,
        RMSE = NA_real_,
        row_type = "predictor",
        stringsAsFactors = FALSE
      ))
    }
  }
  return(detailed_data)
}
```

### Table 1: Full Model Comparison (AICc)
```{r}
#| echo: false
detailed_model_data <- create_detailed_model_table(sorted_models, sorted_delta_aicc, sorted_weights)

# Format for display: replace NA with "" for clean table
display_data <- detailed_model_data %>%
  dplyr::select(-Model_Name) %>%
  mutate(
    across(c(Beta, IRR, Pct_Change, P_Value, Delta_AICc, Weight, RMSE), as.character),
    across(everything(), ~ replace_na(.x, ""))
  ) %>%
  mutate(row_type_flag = row_type) %>%
  dplyr::select(-row_type)

kable(display_data %>% dplyr::select(-row_type_flag),
      col.names = c("Model / Predictor", "β", "IRR", "% Change", "P-Value", "ΔAICc", "w", "RMSE"),
      caption = "Table 1. Negative Binomial Model Comparison (AICc). Models ranked by ΔAICc; coefficients shown as β, IRR, and % Change. Reference: Artificial Refuge, not trail-adjacent, grassland habitat. IRR and % Change are blank for intercept rows, where exp(β) is the baseline rate rather than a ratio.",
      align = c("l", rep("r", 7)),
      digits = 3) %>%
  kable_styling(full_width = FALSE, position = "left",
                bootstrap_options = c("striped", "hover", "condensed")) %>%
  row_spec(which(display_data$row_type_flag == "header"),
           bold = TRUE, background = "#d1ecf1")
```

### Table 2: Competitive Models (ΔAICc < 2)
```{r}
#| echo: false
top_idx <- which(sorted_delta_aicc < 2)
top_models <- sorted_models[top_idx]
top_delta <- sorted_delta_aicc[top_idx]
top_weights <- sorted_weights[top_idx]

cat("=== Competitive Model Set (ΔAICc < 2) ===\n")
cat(sprintf("Number of models within ΔAICc < 2: %d\n", length(top_idx)))
cat(sprintf("Preferred (lowest AICc, highest weight): %s\n\n", best_model_name))

top_model_data <- create_detailed_model_table(top_models, top_delta, top_weights)

top_display <- top_model_data %>%
  dplyr::select(-Model_Name) %>%
  mutate(
    across(c(Beta, IRR, Pct_Change, P_Value, Delta_AICc, Weight, RMSE), as.character),
    across(everything(), ~ replace_na(.x, ""))
  ) %>%
  mutate(row_type_flag = row_type) %>%
  dplyr::select(-row_type)

kable(top_display %>% dplyr::select(-row_type_flag),
      col.names = c("Model / Predictor", "β", "IRR", "% Change", "P-Value", "ΔAICc", "w", "RMSE"),
      caption = sprintf("Table 2. Competitive Models (ΔAICc < 2). Preferred (lowest AICc, highest weight): %s. Reference: Artificial Refuge, not trail-adjacent, grassland habitat. IRR and %% Change are blank for intercept rows, where exp(β) is the baseline rate rather than a ratio.", best_model_name),
      align = c("l", rep("r", 7)),
      digits = 3) %>%
  kable_styling(full_width = FALSE, position = "left",
                bootstrap_options = c("striped", "hover", "condensed")) %>%
  row_spec(which(top_display$row_type_flag == "header"),
           bold = TRUE, background = "#d4edda")
```

### Coefficient Stability Across Competitive Models
```{r}
#| echo: false
# Addresses reviewer's observation that trail/habitat coefficients are unstable
# across top models while feature type is stable (CV comparison)

if (length(top_idx) > 1) {
  coef_names_of_interest <- c(
    "feature_type_methodology_recodedBoulder",
    "feature_type_methodology_recodedLog",
    "trailyes",
    "habitat_typeMarsh",
    "habitat_typeScrub"
  )

  stability_data <- data.frame()
  for (i in seq_along(top_models)) {
    coefs <- coef(top_models[[i]])
    for (cn in coef_names_of_interest) {
      if (cn %in% names(coefs)) {
        stability_data <- rbind(stability_data, data.frame(
          Model = names(top_models)[i],
          Predictor = cn,
          IRR = exp(coefs[cn]),
          stringsAsFactors = FALSE
        ))
      }
    }
  }

  if (nrow(stability_data) > 0) {
    cv_summary <- stability_data %>%
      group_by(Predictor) %>%
      summarise(
        n_models = n(),
        mean_IRR = mean(IRR),
        sd_IRR = sd(IRR),
        CV_pct = ifelse(n() > 1, round(sd(IRR) / mean(IRR) * 100, 1), NA),
        .groups = "drop"
      ) %>%
      mutate(Predictor = case_when(
        Predictor == "feature_type_methodology_recodedBoulder" ~ "Boulder",
        Predictor == "feature_type_methodology_recodedLog" ~ "Log",
        Predictor == "trailyes" ~ "Adjacent to Trail",
        Predictor == "habitat_typeMarsh" ~ "Marsh Habitat",
        Predictor == "habitat_typeScrub" ~ "Scrub Habitat",
        TRUE ~ Predictor
      ))

    cat("=== Coefficient Stability Across Competitive Models ===\n")
    cat("(CV% = coefficient of variation of IRR across models within ΔAICc < 2)\n\n")
    print(cv_summary, digits = 3)
  }
}
```

### DHARMa Diagnostics for All Candidate Models
```{r}
#| echo: false
#| fig-width: 10
#| fig-height: 4
library(DHARMa)

# Diagnostics are assessed BEFORE model selection, so they run over the whole
# candidate set. Restricting them to the ΔAICc < 2 subset would be circular:
# which models are "competitive" is only known after selection.
cat("=== DHARMa Residual Diagnostics for All Candidate Models ===\n\n")

dharma_summary <- data.frame()

for (i in seq_along(sorted_models)) {
  mod_name <- names(sorted_models)[i]
  mod <- sorted_models[[i]]

  cat(sprintf("--- %s (ΔAICc = %.2f, w = %.3f, df = %d)%s ---\n",
              mod_name,
              sorted_delta_aicc[i],
              sorted_weights[i],
              attr(logLik(mod), "df"),
              ifelse(sorted_delta_aicc[i] < 2, "  [competitive]", "")))

  sim_out <- tryCatch(
    simulateResiduals(fittedModel = mod, plot = FALSE, n = 1000),
    error = function(e) { message("DHARMa failed for ", mod_name, ": ", e$message); NULL })

  if (!is.null(sim_out)) {
    dharma_tests <- tryCatch(testResiduals(sim_out, plot = FALSE), error = function(e) NULL)

    if (!is.null(dharma_tests)) {
      ks_p <- dharma_tests$uniformity$p.value
      disp_p <- dharma_tests$dispersion$p.value
      outlier_p <- dharma_tests$outliers$p.value

      cat(sprintf("  KS uniformity test:    p = %.4f %s\n", ks_p, ifelse(ks_p < 0.05, "**", "")))
      cat(sprintf("  Dispersion test:       p = %.4f %s\n", disp_p, ifelse(disp_p < 0.05, "**", "")))
      cat(sprintf("  Outlier test:          p = %.4f %s\n", outlier_p, ifelse(outlier_p < 0.05, "**", "")))

      if (any(c(ks_p, disp_p, outlier_p) < 0.05)) {
        cat("  >> One or more diagnostic tests flagged (p < 0.05). Consider down-weighting this model.\n")
      } else {
        cat("  >> All diagnostics pass.\n")
      }

      dharma_summary <- rbind(dharma_summary, data.frame(
        Model = mod_name, delta_AICc = round(sorted_delta_aicc[i], 2),
        KS = round(ks_p, 4), Dispersion = round(disp_p, 4), Outlier = round(outlier_p, 4),
        Pass = !any(c(ks_p, disp_p, outlier_p) < 0.05), stringsAsFactors = FALSE))
    }
    cat("\n")
  }
}

cat("=== Summary across all candidate models ===\n")
print(dharma_summary, row.names = FALSE)
cat(sprintf("\nModels passing all three tests: %d of %d\n",
            sum(dharma_summary$Pass), nrow(dharma_summary)))
cat(sprintf("Minimum p-value observed across all tests and models: %.4f\n",
            min(c(dharma_summary$KS, dharma_summary$Dispersion, dharma_summary$Outlier))))
```

```{r}
#| echo: false
#| fig-width: 12
#| fig-height: 4
#| fig-cap: "DHARMa residual diagnostics for all competitive models (ΔAICc < 2). QQ plot and residuals vs. predicted for each model."

n_top <- length(top_models)
par(mfrow = c(1, n_top * 2), mar = c(4, 3, 3, 1))

for (i in seq_along(top_models)) {
  mod_name <- names(top_models)[i]
  mod <- top_models[[i]]
  sim_out <- tryCatch(
    simulateResiduals(fittedModel = mod, plot = FALSE, n = 1000),
    error = function(e) NULL)

  if (!is.null(sim_out)) {
    plotQQunif(sim_out, main = paste0(mod_name, "\nQQ Plot"))
    plotResiduals(sim_out, main = paste0(mod_name, "\nResid vs Pred"))
  }
}

par(mfrow = c(1, 1))
```

### Sensitivity Analysis: Interior Sites Only
```{r}
#| echo: false
# ============================================================================
# Exclude the 5 trail-adjacent (edge) sites and rerun model comparison.
# Addresses reviewer concern that periphery sites may bias results.
# ============================================================================

interior_data <- placename_summary_pt_2 %>%
  filter(trail == "no")

cat(sprintf("Interior-only dataset: n = %d sites (excluded %d trail-adjacent sites)\n",
            nrow(interior_data),
            nrow(placename_summary_pt_2) - nrow(interior_data)))
cat("\nFeature type breakdown:\n")
interior_data %>% count(feature_type_methodology_recoded) %>% print()
cat("\nHabitat type breakdown:\n")
interior_data %>% count(habitat_type) %>% print()

# --- Fit models on interior sites only (trail variable dropped since all are interior) ---

nb_int_feat <- tryCatch(
  MASS::glm.nb(total_detections ~ feature_type_methodology_recoded + offset(log(total_camera_hours)),
               data = interior_data),
  error = function(e) { message("Feature Type failed: ", e$message); NULL })

nb_int_hab <- tryCatch(
  MASS::glm.nb(total_detections ~ habitat_type + offset(log(total_camera_hours)),
               data = interior_data),
  error = function(e) { message("Habitat Type failed: ", e$message); NULL })

nb_int_feat_hab <- tryCatch(
  MASS::glm.nb(total_detections ~ feature_type_methodology_recoded + habitat_type + offset(log(total_camera_hours)),
               data = interior_data),
  error = function(e) { message("Feature + Habitat failed: ", e$message); NULL })

nb_int_feat_x_hab <- tryCatch(
  MASS::glm.nb(total_detections ~ feature_type_methodology_recoded * habitat_type + offset(log(total_camera_hours)),
               data = interior_data),
  error = function(e) { message("Feature x Habitat interaction failed: ", e$message); NULL })

# Intercept-only (null) model for reference
nb_int_null <- tryCatch(
  MASS::glm.nb(total_detections ~ 1 + offset(log(total_camera_hours)),
               data = interior_data),
  error = function(e) { message("Null model failed: ", e$message); NULL })

# Build model list
interior_models <- list(
  "Null (intercept only)" = nb_int_null,
  "Feature Type" = nb_int_feat,
  "Habitat Type" = nb_int_hab,
  "Feature Type + Habitat Type" = nb_int_feat_hab,
  "Feature Type x Habitat Type" = nb_int_feat_x_hab
)
# Drop models that failed to converge or are rank deficient (see is_estimable()).
interior_models <- interior_models[sapply(interior_models, is_estimable)]

# AICc comparison
int_aicc <- data.frame(
  Model = names(interior_models),
  df = sapply(interior_models, function(m) attr(logLik(m), "df")),
  AICc = sapply(interior_models, MuMIn::AICc),
  stringsAsFactors = FALSE
)
int_aicc$delta_AICc <- int_aicc$AICc - min(int_aicc$AICc)
int_aicc$rel_likelihood <- exp(-0.5 * int_aicc$delta_AICc)
int_aicc$akaike_weight <- int_aicc$rel_likelihood / sum(int_aicc$rel_likelihood)
int_aicc <- int_aicc %>%
  arrange(delta_AICc) %>%
  dplyr::select(Model, df, AICc, delta_AICc, akaike_weight)

cat("\n=== Interior-Only AICc Model Comparison (n =", nrow(interior_data), ") ===\n")
print(int_aicc, digits = 3, row.names = FALSE)

# Preferred model (lowest AICc, highest weight)
int_competitive <- which(int_aicc$delta_AICc < 2)
int_best_idx <- int_competitive[1]
int_best_name <- int_aicc$Model[int_best_idx]
int_best_model <- interior_models[[int_best_name]]

cat(sprintf("\nPreferred model (lowest AICc, highest weight): %s (w = %.3f)\n",
    int_best_name, int_aicc$akaike_weight[int_best_idx]))
cat("\nBest model summary:\n")
summary(int_best_model)
```

```{r}
#| echo: false
# Detailed table for interior-only models
int_model_order <- match(int_aicc$Model, names(interior_models))
int_sorted_models <- interior_models[int_model_order]
int_sorted_delta <- int_aicc$delta_AICc
int_sorted_weights <- int_aicc$akaike_weight

int_table_data <- create_detailed_model_table(int_sorted_models, int_sorted_delta, int_sorted_weights,
                                               observed = interior_data$total_detections)

int_display <- int_table_data %>%
  dplyr::select(-Model_Name) %>%
  mutate(
    across(c(Beta, IRR, Pct_Change, P_Value, Delta_AICc, Weight, RMSE), as.character),
    across(everything(), ~ replace_na(.x, ""))
  ) %>%
  mutate(row_type_flag = row_type) %>%
  dplyr::select(-row_type)

kable(int_display %>% dplyr::select(-row_type_flag),
      col.names = c("Model / Predictor", "β", "IRR", "% Change", "P-Value", "ΔAICc", "w", "RMSE"),
      caption = sprintf("Table 3. Interior Sites Only (n=%d). Models ranked by ΔAICc. Reference: Artificial Refuge, grassland habitat. IRR and %% Change are blank for intercept rows, where exp(β) is the baseline rate rather than a ratio.", nrow(interior_data)),
      align = c("l", rep("r", 7)),
      digits = 3) %>%
  kable_styling(full_width = FALSE, position = "left",
                bootstrap_options = c("striped", "hover", "condensed")) %>%
  row_spec(which(int_display$row_type_flag == "header"),
           bold = TRUE, background = "#fce4ec")
```

```{r}
#| echo: false
# Moran's I on interior-only best model
int_coords <- interior_data %>%
  dplyr::select(longitude, latitude) %>%
  as.matrix()

if (nrow(int_coords) >= 5) {
  int_knn <- spdep::knearneigh(int_coords, k = min(5, nrow(int_coords) - 1))
  int_nb <- spdep::knn2nb(int_knn)
  int_listw <- spdep::nb2listw(int_nb, style = "W")
  int_resids <- residuals(int_best_model, type = "deviance")
  int_moran <- spdep::moran.test(int_resids, int_listw)

  cat("=== Moran's I on Interior-Only Best Model ===\n")
  cat(sprintf("Model: %s, n = %d\n", int_best_name, nrow(interior_data)))
  print(int_moran)

  if (int_moran$p.value < 0.05) {
    cat("\n** Significant spatial autocorrelation detected in interior subset. **\n")
  } else {
    cat("\n** No significant spatial autocorrelation in interior subset. **\n")
  }
}
```


## Figures

### Daily Detections Mean and 95% Confidence Intervals by Feature Type
```{r fig-daily-detections-mean-ci}
#| fig-cap: "Average wildlife detections per camera trap day across all three feature type groups—artificial refuges, boulders, and logs. Points represent individual sites, black diamonds indicate mean values, and vertical lines show 95% confidence intervals."

# Step 1: Compute summary stats
summary_pt_2_by_feature_type_recoded <- placename_summary_pt_2 %>%
  group_by(feature_type_methodology_recoded) %>%
  summarise(
    mean = mean(detections_per_camera_day, na.rm = TRUE),
    se = sd(detections_per_camera_day, na.rm = TRUE) / sqrt(n()),
    n = n(),
    .groups = "drop"
  ) %>%
  mutate(
    lower = mean - qt(0.975, df = n - 1) * se,
    upper = mean + qt(0.975, df = n - 1) * se
  )

# Step 2: Plot
p1 = ggplot(placename_summary_pt_2, aes(x = feature_type_methodology_recoded, y = detections_per_camera_day)) +
  geom_jitter(aes(color = feature_type_methodology_recoded), width = 0.25, alpha = 0.6, size = 3.5, show.legend = FALSE) +
  geom_errorbar(data = summary_pt_2_by_feature_type_recoded, aes(x = feature_type_methodology_recoded, y = mean, ymin = lower, ymax = upper),
                width = 0.15, color = "black", linewidth = 1) +
  geom_point(data = summary_pt_2_by_feature_type_recoded,
             aes(x = feature_type_methodology_recoded, y = mean),
             shape = 23, size = 6, fill = "black", color = "black", stroke = 1.5) +
  scale_color_manual(values = c("Boulder" = "#66c2a5", "Log" = "#8da0cb", "Artificial Refuge" = "#fc8d62")) +
  labs(
    title = "Detection Rate per Camera Day by Feature Type",
    x = "Feature Type",
    y = "Detections per camera day"
  ) +
  theme_cowplot() +
  theme(
    axis.text.x = element_text(vjust = 1, size = 14),
    axis.text.y = element_text(size = 12),
    axis.title.x = element_text(size = 14),
    axis.title.y = element_text(size = 14),
    plot.title = element_text(size = 16, face = "bold"),
    legend.position = "none"
  )

# Save the plot as a PNG (narrower width)
ggsave(here("figures","feature_type_detection_rate_pt_2.png"), plot = p1, width = 6, height = 6, dpi = 300, bg = "white")
p1
```

### Species by Habitat Type Analysis
```{r fig-species-by-habitat-type}
#| fig-cap: "Species detection rates per camera day by habitat type."
#| fig-pos: 'H'

### Species Lumping
# Set detection threshold (below which species are lumped by taxonomic class)
detection_threshold <- 50

# Count total observations by species
species_counts_pt_2 <- clean_pt_2_seq_dep %>%
  filter(!is.na(common_name)) %>%
  count(common_name, name = "total_count")

# Recode class names for grouping
clean_pt_2_seq_dep_lumped <- clean_pt_2_seq_dep %>%
  mutate(
    class_grouped = case_when(
      class == "Aves" ~ "Birds",
      class == "Mammalia" ~ "Mammals",
      class == "Reptilia" ~ "Reptiles",
      TRUE ~ class
    )
  ) %>%
  left_join(species_counts_pt_2, by = "common_name") %>%
  mutate(
    common_name_lumped = case_when(
      total_count < detection_threshold & !is.na(class_grouped) ~ paste0("Other (", class_grouped, ")"),
      grepl("other|unidentified", tolower(common_name)) & !is.na(class_grouped) ~ paste0("Other (", class_grouped, ")"),
      TRUE ~ common_name
    )
  )

# Build species palette
lumped_species_pt_2 <- clean_pt_2_seq_dep_lumped %>%
  filter(!is.na(common_name_lumped)) %>%
  distinct(common_name_lumped) %>%
  mutate(
    is_other = grepl("^Other", common_name_lumped),
    sort_key = ifelse(is_other, paste0("zzz_", common_name_lumped), common_name_lumped)
  ) %>%
  arrange(sort_key) %>%
  pull(common_name_lumped)

species_palette_pt_2 <- setNames(
  colorRampPalette(RColorBrewer::brewer.pal(8, "Set2"))(length(lumped_species_pt_2)),
  lumped_species_pt_2
)

# summarize effort by habitat type
habitat_effort_summary_pt_2 <- clean_pt_2_dep %>%
  group_by(habitat_type) %>%
  summarise(
    total_camera_days = sum(deployment_duration, na.rm = TRUE),
    total_camera_hours = total_camera_days * 24,
    .groups = "drop"
  )

habitat_detections_summary_pt_2 <- clean_pt_2_seq_dep %>%
  group_by(habitat_type) %>%
  summarise(
    total_habitat_detections = sum(group_size, na.rm = TRUE),
    .groups = "drop"
  )

# Join effort and detections by habitat type
habitat_summary_pt_2 <- habitat_effort_summary_pt_2 |>
  dplyr::left_join(habitat_detections_summary_pt_2, by = "habitat_type") |>
  dplyr::mutate(
    habitat_detections_per_camera_day = total_habitat_detections / total_camera_days,
    habitat_detections_per_camera_hour = total_habitat_detections / total_camera_hours
  )

# Prepare data for plot: 
detections_by_habitat_species_pt_2 <- clean_pt_2_seq_dep_lumped %>%
  group_by(habitat_type, common_name_lumped) %>%
  summarise(total_species_detections_in_habitat = sum(group_size, na.rm = TRUE), .groups = "drop") %>%
  arrange(habitat_type, desc(total_species_detections_in_habitat))

detections_by_habitat_species_pt_2 <- clean_pt_2_seq_dep_lumped %>%
  group_by(habitat_type, common_name_lumped) %>%
  summarise(total_species_detections_in_habitat = sum(group_size, na.rm = TRUE), .groups = "drop") %>%
  arrange(habitat_type, desc(total_species_detections_in_habitat)) %>%
  left_join(habitat_effort_summary_pt_2, by = "habitat_type") |> 
  mutate(
  species_detections_per_camera_day = total_species_detections_in_habitat / total_camera_days,
  species_detections_per_camera_hour = total_species_detections_in_habitat / total_camera_hours
)

# Set factor levels to order legend manually
detections_by_habitat_species_pt_2$common_name_lumped <- factor(detections_by_habitat_species_pt_2$common_name_lumped, levels = c(
  "Brush Rabbit",
  "California Ground Squirrel",
  "North American Deermouse",
  "Western fence lizard",
  "Other (Birds)",
  "Other (Mammals)",
  "Other (Reptiles)"
))

##stacked bar chart
p2 = ggplot(detections_by_habitat_species_pt_2 %>%
              filter(common_name_lumped != "Other (Reptiles)"),
       aes(x = habitat_type,
           y = species_detections_per_camera_day,
           fill = common_name_lumped)) +
  geom_bar(stat = "identity") +
  labs(
    x = "Habitat Type",
    y = "Detections per Camera Day",
    fill = "Species",
    title = ""
  ) +
  scale_fill_manual(values = species_palette_pt_2) +
  scale_y_continuous(breaks = seq(0, 10, by = 2), expand = expansion(mult = c(0, 0.05))) +
  theme_minimal(base_family = "sans") +
  theme(
    plot.title = element_text(size = 14, face = "bold"),
    panel.grid.major.x = element_blank()
  )
p2 <- p2 + theme_cowplot() +
  theme(
    legend.text = element_text(size = 11),
    axis.title = element_text(size = 14),
    axis.text = element_text(size = 12)
  )

# Save the plot as a PNG
ggsave(here("figures","habitat_type_species_detection_rates_pt_2.png"), plot = p2, width = 8, height = 6, dpi = 300, bg = "white")
p2

```

### Four Resident Species by Habitat Type (4-panel)
```{r fig-4panel-species-habitat}
#| fig-cap: "Detection rates per camera day for the four resident species across habitat types. Each panel shows one species; bars are colored by habitat type."
#| fig-width: 10
#| fig-height: 8

resident_species <- c("Brush Rabbit", "California Ground Squirrel",
                       "North American Deermouse", "Western fence lizard")

habitat_colors <- c("Marsh" = "#1b9e77", "Grassland" = "#d95f02", "Scrub" = "#7570b3")

resident_habitat_data <- clean_pt_2_seq_dep %>%
  filter(common_name %in% resident_species) %>%
  group_by(common_name, habitat_type) %>%
  summarise(total_detections = sum(group_size, na.rm = TRUE), .groups = "drop") %>%
  left_join(habitat_effort_summary_pt_2, by = "habitat_type") %>%
  mutate(detections_per_camera_day = total_detections / total_camera_days)

# Ensure all species x habitat combos exist (fill zeros)
resident_habitat_data <- resident_habitat_data %>%
  complete(common_name = resident_species,
           habitat_type = c("Marsh", "Grassland", "Scrub"),
           fill = list(total_detections = 0, detections_per_camera_day = 0))

p_4panel <- ggplot(resident_habitat_data,
       aes(x = habitat_type, y = detections_per_camera_day, fill = habitat_type)) +
  geom_col(width = 0.7) +
  facet_wrap(~ common_name, scales = "free_y", ncol = 2) +
  scale_fill_manual(values = habitat_colors) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.1))) +
  labs(
    x = "Habitat Type",
    y = "Detections per Camera Day",
    fill = "Habitat"
  ) +
  theme_cowplot() +
  theme(
    strip.text = element_text(size = 12, face = "italic"),
    axis.text = element_text(size = 11),
    axis.title = element_text(size = 13),
    legend.position = "bottom",
    legend.text = element_text(size = 11)
  )

ggsave(here("figures", "4panel_resident_species_by_habitat.png"),
       plot = p_4panel, width = 10, height = 8, dpi = 300, bg = "white")
p_4panel
```

### Hourly Species Detections by Feature Type

```{r}
# summarize effort by habitat type
feature_type_effort_summary_pt_2 <- clean_pt_2_dep %>%
  group_by(feature_type_methodology) %>%
  summarise(
    total_camera_days = sum(deployment_duration, na.rm = TRUE),
    total_camera_hours = total_camera_days * 24,
    .groups = "drop"
  )

# Define all hour levels and updated species levels
levels_hours <- c("12 AM", sprintf("%02d AM", 1:11), "12 PM", sprintf("%02d PM", 1:11))
levels_species <- lumped_species_pt_2

# Helper to pad missing combinations by feature type
pad_hourly <- function(df, feature_type) {
  df %>%
    filter(feature_type_methodology == feature_type, !is.na(common_name_lumped)) %>%
    mutate(
      hour_am_pm = format(start_time, "%I %p"),
      hour_am_pm = factor(hour_am_pm, levels = levels_hours),
      common_name_lumped = factor(common_name_lumped, levels = levels_species)
    ) %>%
    count(hour_am_pm, common_name_lumped, name = "detections") %>%
    complete(hour_am_pm, common_name_lumped, fill = list(count = 0))
}

# Create padded datasets
# Boulders
figs_hourly_boulder <- pad_hourly(clean_pt_2_seq_dep_lumped, "Boulder")
# Step 1: Extract camera days for Boulder
boulder_effort <- feature_type_effort_summary_pt_2 |>
  dplyr::filter(feature_type_methodology == "Boulder") |>
  dplyr::pull(total_camera_days)
# Step 2: Normalize the counts
figs_hourly_boulder <- figs_hourly_boulder |>
  dplyr::mutate(detections_per_camera_day = detections / boulder_effort)

# Artificial Refuge (AR)
figs_hourly_hibernacula <- pad_hourly(clean_pt_2_seq_dep_lumped, "Artificial Refuge")
hibernacula_effort <- feature_type_effort_summary_pt_2 |>
  dplyr::filter(feature_type_methodology == "Artificial Refuge") |>
  dplyr::pull(total_camera_days)
figs_hourly_hibernacula <- figs_hourly_hibernacula |>
  dplyr::mutate(detections_per_camera_day = detections / hibernacula_effort)

# Logs
figs_hourly_log <- pad_hourly(clean_pt_2_seq_dep_lumped, "Log")
log_effort <- feature_type_effort_summary_pt_2 |>
  dplyr::filter(feature_type_methodology == "Log") |>
  dplyr::pull(total_camera_days)
figs_hourly_log <- figs_hourly_log |>
  dplyr::mutate(detections_per_camera_day = detections / log_effort)

# Reusable theme
shared_theme <- theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)
  )

# Set consistent y-axis max
max_count <- 1

# Plot for Boulder
figs_plot_boulder <- ggplot(figs_hourly_boulder, aes(x = hour_am_pm, y = detections_per_camera_day, fill = common_name_lumped)) +
  geom_bar(stat = "identity", position = "stack") +
  scale_fill_manual(values = species_palette_pt_2, name = "Species") +
  scale_y_continuous(limits = c(0, max_count)) +
  labs(
    title = "Boulder Sites",
    x = "Hour of Day (AM/PM)",
    y = ""
  ) +
  shared_theme

# Plot for Artificial Refuge (AR)
figs_plot_hibernacula <- ggplot(figs_hourly_hibernacula, aes(x = hour_am_pm, y = detections_per_camera_day, fill = common_name_lumped)) +
  geom_bar(stat = "identity", position = "stack") +
  scale_fill_manual(values = species_palette_pt_2, name = "Species") +
  scale_y_continuous(limits = c(0, max_count)) +
  labs(
    title = "Artificial Refuge",
    x = "",
    y = "Detections Per Camera Day"
  ) +
  shared_theme

# Plot for Log
figs_plot_log <- ggplot(figs_hourly_log, aes(x = hour_am_pm, y = detections_per_camera_day, fill = common_name_lumped)) +
  geom_bar(stat = "identity", position = "stack") +
  scale_fill_manual(values = species_palette_pt_2, name = "Species") +
  scale_y_continuous(limits = c(0, max_count)) +
  labs(
    title = "Log Sites",
    x = "",
    y = ""
  ) +
  shared_theme

# Combine plots using patchwork
p3 = (figs_plot_hibernacula | figs_plot_boulder | figs_plot_log) +
  plot_layout(guides = "collect") &
  theme(legend.position = "bottom")

p3

ggsave(here("figures","diel_species_detection_rates_by_feature_type_pt_2.png"), plot = p3, width = 12, height = 6, dpi = 300, bg = "white")
```

### Species List
```{r tbl-species-list}
#| tbl-cap: "Species List for Part 2 of the Study. Only includes wildlife identified to species level. Table is sorted by class then total number of detections." 

# Calculate total unique features (placename) per feature_type
total_features_pt_2 <- clean_pt_2_seq_dep %>%
  distinct(feature_type_methodology, placename) %>%
  group_by(feature_type_methodology) %>%
  summarise(total_features_pt_2 = n(), .groups = "drop")

# Summarize sum(group_size) and count unique visited placename per species per feature type
species_summary_pt_2 <- clean_pt_2_seq_dep %>%
  group_by(class, common_name, genus_species, feature_type_methodology) %>%
  summarise(
    count_observed = sum(group_size, na.rm = TRUE),
    features_visited = n_distinct(placename),
    .groups = "drop"
  )

# Pivot counts wide
species_counts_wide_pt_2 <- species_summary_pt_2 %>%
  dplyr::select(-features_visited) %>%
  pivot_wider(
    names_from = feature_type_methodology,
    values_from = count_observed,
    values_fill = 0,
    names_prefix = "count_observed_at_"
  ) |> 
  clean_names()

# Pivot features visited wide
species_features_wide_pt_2 <- species_summary_pt_2 %>%
  dplyr::select(class, common_name, genus_species, feature_type_methodology, features_visited) %>%
  pivot_wider(
    names_from = feature_type_methodology,
    values_from = features_visited,
    values_fill = 0,
    names_prefix = "features_visited_at_"
  ) |> 
  clean_names()

# Combine counts and features visited
species_combined_pt_2 <- species_counts_wide_pt_2 %>%
  full_join(species_features_wide_pt_2, by = c("class", "common_name", "genus_species"))

# Extract total features per feature type as named vector
total_features_pt_2_vec <- total_features_pt_2 %>%
  deframe()  # named vector: names are feature_type_methodology, values total_features_pt_2

# Calculate percent visited by species at each feature type
species_final_pt_2 <- species_combined_pt_2 %>%
  mutate(
    percent_hibernacula_visited = 100 * features_visited_at_artificial_refuge / total_features_pt_2_vec["Artificial Refuge"],
    percent_boulders_visited = 100 * features_visited_at_boulder / total_features_pt_2_vec["Boulder"],
    percent_logs_visited = 100 * features_visited_at_log / total_features_pt_2_vec["Log"]
  )

camera_days_vec <- feature_type_effort_summary_pt_2 %>%
  dplyr::select(feature_type_methodology, total_camera_days) %>%
  pull(total_camera_days) %>%
  setNames(feature_type_effort_summary_pt_2$feature_type_methodology)

total_camera_days_all <- sum(camera_days_vec)

species_final_pt_2 <- species_final_pt_2 %>%
  mutate(
    rate_per_cam_day_hibernacula = count_observed_at_artificial_refuge / camera_days_vec["Artificial Refuge"],
    rate_per_cam_day_boulder = count_observed_at_boulder / camera_days_vec["Boulder"],
    rate_per_cam_day_log = count_observed_at_log / camera_days_vec["Log"]
  )

t1 = species_final_pt_2 %>%
  # Select and rename columns for a cleaner table
  dplyr::select(
    'Class' = class,
    'Common Name' = common_name,
    'Scientific Name' = genus_species,
    'Detections' = count_observed_at_artificial_refuge,
    '% Visited' = percent_hibernacula_visited,
    'Per Cam.-Day' = rate_per_cam_day_hibernacula,
    'Detections ' = count_observed_at_boulder,
    '% Visited ' = percent_boulders_visited,
    'Per Cam.-Day ' = rate_per_cam_day_boulder,
    'Detections  ' = count_observed_at_log,
    '% Visited  ' = percent_logs_visited,
    'Per Cam.-Day  ' = rate_per_cam_day_log
  ) %>%
  mutate(
    'Detections   ' = Detections + `Detections ` + `Detections  `,
    '% Visited   ' = (
      (`% Visited` * total_features_pt_2_vec["Artificial Refuge"]) +
      (`% Visited ` * total_features_pt_2_vec["Boulder"]) +
      (`% Visited  ` * total_features_pt_2_vec["Log"])
    ) / sum(total_features_pt_2_vec),
    'Per Cam.-Day   ' = ((Detections + `Detections ` + `Detections  `) / total_camera_days_all),
    # Round and format
    across(starts_with("%"), ~ paste0(round(.x, 2), "%")),
    across(starts_with("Rate/Day"), ~ round(.x, 2))
  ) %>%
  arrange(Class, desc(`Detections   `), `Common Name`) %>%
  kable(
    caption = "Species Observations and Percent Feature Visited by Feature Type",
    align = c("l", "l", "l", rep("c", 12)),
    digits = 3
  ) %>%
  kable_styling(
    full_width = FALSE,
    position = "left",
    bootstrap_options = c("striped", "hover", "condensed")
  ) %>%
  add_header_above(c(" " = 3, "Artificial Refuges" = 3, "Boulders" = 3, "Logs" = 3, "Total" = 3))

# Save the kable as an HTML file
tryCatch({
  save_kable(t1, here("misc","species_table_pt_2.html"))
  webshot(here("misc","species_table_pt_2.html"), here("figures","species_table_pt_2.png"), zoom = 2, vwidth = 1800)
}, error = function(e) message("Table save/screenshot skipped: ", e$message))
```

### Map
```{r fig-deployment-map}
#| fig-cap: "**Spatial distribution of camera trap deployments at the study site.** Circle color indicates feature type and size reflects the average daily number of wildlife observations. Inset map shows the location within California."
#| fig-width: 10
#| fig-height: 6

# Convert site_summary to sf object
map_site_sf <- st_as_sf(placename_summary_pt_2, coords = c("longitude", "latitude"), crs = 4326)

# Load NCOS habitat layer from local shapefile
ncos_habitats <- st_read(here("data", "ncos_habitats_june_2021", "NCOS_Habitats_June2021_Simple.shp"))

# Fix invalid geometries and transform to match CRS
ncos_habitats <- st_make_valid(ncos_habitats)
ncos_habitats <- st_transform(ncos_habitats, crs = st_crs(map_site_sf))

# Simplify habitat categories and dissolve by group
ncos_habitats <- ncos_habitats |>
  mutate(habitat_simple = case_when(
    str_detect(Habitat, "Grassland") ~ "Grassland",
    str_detect(Habitat, "Marsh|marsh") ~ "Marsh",
    str_detect(Habitat, "Scrub") ~ "Scrub",
    TRUE ~ "Other"
  )) |>
  group_by(habitat_simple) |>
  summarise(geometry = st_union(geometry), .groups = "drop")

# Load NCOS boundary and convert to lines (so clipping doesn't create false polygon edges)
ncos_boundary <- st_read(here("data", "ncos_shp", "ncos_shp.shp"))
ncos_boundary <- st_make_valid(ncos_boundary)
ncos_boundary <- st_transform(ncos_boundary, crs = st_crs(map_site_sf))
ncos_boundary_lines <- st_cast(ncos_boundary, "MULTILINESTRING")

# Get bounding box and expand slightly
bbox <- st_bbox(map_site_sf)
padding <- 0.01  # Reduced padding for tighter view
west_shift <- 0.0  # Adjust this value to shift more/less west
bbox_expanded <- c(
  xmin = bbox["xmin"] - padding - west_shift,
  xmax = bbox["xmax"] + padding - west_shift,
  ymin = bbox["ymin"] - padding,
  ymax = bbox["ymax"] + padding
)

# Download satellite tiles (Esri)
map_basemap <- get_tiles(map_site_sf, provider = "Esri.WorldImagery", zoom = 18)

# Clip habitat layer and boundary lines to satellite basemap extent
ncos_habitats <- st_crop(ncos_habitats, st_bbox(map_basemap))
ncos_boundary_lines <- st_crop(ncos_boundary_lines, st_bbox(map_basemap))

# Convert raster to data.frame for ggplot
map_rgb_df <- as.data.frame(map_basemap, xy = TRUE)
colnames(map_rgb_df) <- c("x", "y", "red", "green", "blue")
map_rgb_df$hex <- rgb(map_rgb_df$red, map_rgb_df$green, map_rgb_df$blue, maxColorValue = 255)

# Set feature type colors
feature_colors <- c(
  "Boulder" = brewer.pal(3, "Set2")[1],
  "Artificial Refuge" = brewer.pal(3, "Set2")[2],
  "Log" = brewer.pal(3, "Set2")[3]
)

# Define simplified habitat colors (excluding "Other")
habitat_colors <- c(
  "Grassland" = "#9ACD32",
  "Marsh" = "#2E8B57",
  "Scrub" = "#9370DB"
)

# Filter out "Other" habitat type
ncos_habitats_filtered <- ncos_habitats |> filter(habitat_simple != "Other")

# Main satellite deployment map with habitat layer
map_main <- ggplot() +
  geom_raster(data = map_rgb_df, aes(x = x, y = y, fill = hex)) +
  scale_fill_identity() +
  # Add NCOS boundary (yellow outline, no fill)
  geom_sf(data = ncos_boundary_lines |> mutate(boundary = "NCOS Boundary"),
          aes(linetype = boundary), color = "yellow", linewidth = 1) +
  scale_linetype_manual(values = "solid", name = NULL) +
  # Add habitat polygons (semi-transparent overlay on satellite)
  ggnewscale::new_scale_fill() +
  geom_sf(data = ncos_habitats_filtered, aes(fill = habitat_simple), color = NA, alpha = 0.5) +
  scale_fill_manual(values = habitat_colors, name = "Habitat Type") +
  # Add camera deployment points
  ggnewscale::new_scale_fill() +
  geom_sf(data = map_site_sf, aes(fill = feature_type_methodology, size = detections_per_camera_day),
          shape = 21, color = "black", stroke = 0.8, alpha = 0.9) +
  scale_fill_manual(values = feature_colors, name = "Feature Type") +
  scale_size(name = "Avg. Daily Obs.", range = c(2, 6)) +
  coord_sf(xlim = c(bbox_expanded["xmin"], bbox_expanded["xmax"]),
           ylim = c(bbox_expanded["ymin"], bbox_expanded["ymax"]),
           clip = "on") +
  annotation_scale(
    location = "bl",      # bottom left
    pad_x = unit(0.45, "in"),
    pad_y = unit(0.3, "in"),
    width_hint = 0.25,
    text_col = "white"   # White text
  ) +
  annotation_north_arrow(
    location = "br",      # bottom right
    which_north = "true",
    pad_x = unit(0.35, "in"),
    pad_y = unit(0.23, "in"),
    style = north_arrow_fancy_orienteering
  ) +
  theme_minimal(base_size = 12) +
  theme(
    panel.grid = element_blank(),
    legend.position = "bottom",
    axis.title = element_blank(),
    axis.text = element_blank(),
    axis.ticks = element_blank(),
    legend.box = "vertical",
    legend.box.spacing = unit(0, "cm"),
    legend.spacing.y = unit(0, "cm"),
    legend.margin = margin(t = 0, b = 0),
    legend.title = element_text(size = 10),
    legend.text = element_text(size = 9),
    legend.key.height = unit(0.4, "cm")
  ) +
  guides(
    linetype = guide_legend(order = 1, override.aes = list(color = "yellow", linewidth = 1)),
    fill = guide_legend(order = 2, nrow = 1, override.aes = list(alpha = 0.9, size = 4, shape = 21, stroke = 0.8)),
    size = guide_legend(order = 3, nrow = 1, override.aes = list(fill = "black", shape = 21, stroke = 0.8))
  )

# Calculate study area centroid
map_centroid_coords <- map_site_sf %>%
  st_coordinates() %>%
  as.data.frame() %>%
  summarise(longitude = mean(X), latitude = mean(Y))

# Overview map of CA with study area location
map_CA <- map("state", plot = FALSE, fill = TRUE) %>%
  st_as_sf() %>%
  filter(ID == "california")

map_overview <- ggplot() +
  geom_sf(data = map_CA, fill = "gray90", color = "gray60", size = 0.3) +
  geom_point(data = map_centroid_coords, aes(x = longitude, y = latitude), 
             color = "red", size = 3, shape = 15) +
  theme_void() +
  theme(
    panel.border = element_rect(color = "black", fill = NA, size = 0.5),
    plot.background = element_rect(fill = "white", color = NA)
  )

# Combine maps with proper positioning
map_final <- ggdraw() +
  draw_plot(map_main) +
  draw_plot(map_overview, x = 0.68, y = 0.647, width = 0.3, height = 0.3)

map_final

ggsave(here("figures","deployment_map_pt_2.png"), plot = map_final, width = 10, height = 6, dpi = 300, bg = "white")
```
### Map: Spring Study (Full NCOS Boundary)
```{r fig-deployment-map-2}
#| fig-cap: "**Spring study: spatial distribution of camera trap deployments.** Point shape indicates feature type; point size reflects average daily wildlife detections. Habitat polygons shown as semi-transparent overlays."
#| fig-width: 10
#| fig-height: 6

# Convert site summary to sf
map_site_sf <- st_as_sf(placename_summary_pt_2, coords = c("longitude", "latitude"), crs = 4326)

# Load and process habitat layer
ncos_habitats <- st_read(here("data", "ncos_habitats_june_2021", "NCOS_Habitats_June2021_Simple.shp"))
ncos_habitats <- st_make_valid(ncos_habitats)
ncos_habitats <- st_transform(ncos_habitats, crs = st_crs(map_site_sf))
ncos_habitats <- ncos_habitats |>
  mutate(habitat_simple = case_when(
    str_detect(Habitat, "Grassland") ~ "Grassland",
    Habitat == "Upland Mosaic" ~ "Grassland",
    str_detect(Habitat, "Marsh|marsh") ~ "Marsh",
    str_detect(Habitat, "Scrub") ~ "Scrub",
    Habitat == "Oak Woodland/Chaparral" ~ "Scrub",
    TRUE ~ "Other"
  )) |>
  group_by(habitat_simple) |>
  summarise(geometry = st_union(geometry), .groups = "drop")

# Load NCOS boundary
ncos_boundary <- st_read(here("data", "ncos_shp", "ncos_shp.shp"))
ncos_boundary <- st_make_valid(ncos_boundary)
ncos_boundary <- st_transform(ncos_boundary, crs = st_crs(map_site_sf))
ncos_boundary_lines <- st_cast(ncos_boundary, "MULTILINESTRING")

# Bbox from NCOS boundary (not site points)
bbox_expanded <- st_bbox(ncos_boundary)
bbox_sf <- st_as_sfc(bbox_expanded)

# Download satellite tiles covering full boundary
map_basemap <- get_tiles(bbox_sf, provider = "Esri.WorldImagery", zoom = 18)

# Clip layers to basemap extent
ncos_habitats <- st_crop(ncos_habitats, st_bbox(map_basemap))
ncos_boundary_lines <- st_crop(ncos_boundary_lines, st_bbox(map_basemap))

# Convert raster for ggplot
map_rgb_df <- as.data.frame(map_basemap, xy = TRUE)
colnames(map_rgb_df) <- c("x", "y", "red", "green", "blue")
map_rgb_df$hex <- rgb(map_rgb_df$red, map_rgb_df$green, map_rgb_df$blue, maxColorValue = 255)

habitat_colors <- c("Grassland" = "#9ACD32", "Marsh" = "#2E8B57", "Scrub" = "#9370DB")
ncos_habitats_filtered <- ncos_habitats |> filter(habitat_simple != "Other")

# Shape encoding: square=refugium, triangle=boulder, circle=log
point_shapes_pt2 <- c("Artificial Refuge" = 15, "Boulder" = 2, "Log" = 1)

map_main <- ggplot() +
  geom_raster(data = map_rgb_df, aes(x = x, y = y, fill = hex)) +
  scale_fill_identity() +
  geom_sf(data = ncos_boundary_lines |> mutate(boundary = "NCOS Boundary"),
          aes(linetype = boundary), color = "yellow", linewidth = 1) +
  scale_linetype_manual(values = "solid", name = NULL) +
  ggnewscale::new_scale_fill() +
  geom_sf(data = ncos_habitats_filtered, aes(fill = habitat_simple), color = NA, alpha = 0.5) +
  scale_fill_manual(values = habitat_colors, name = "Habitat Type") +
  geom_sf(data = map_site_sf,
          aes(shape = feature_type_methodology, size = detections_per_camera_day),
          color = "black", stroke = 1.2) +
  scale_shape_manual(values = point_shapes_pt2, name = "Feature Type") +
  scale_size(name = "Avg. Daily Obs.", range = c(2, 6)) +
  coord_sf(expand = TRUE) +
  annotation_scale(location = "bl", pad_x = unit(0.45, "in"), pad_y = unit(0.3, "in"),
                   width_hint = 0.25, text_col = "white") +
  annotation_north_arrow(location = "br", which_north = "true",
                         pad_x = unit(0.35, "in"), pad_y = unit(0.23, "in"),
                         style = north_arrow_fancy_orienteering) +
  theme_minimal(base_size = 12) +
  theme(
    panel.grid = element_blank(),
    legend.position = "bottom",
    axis.title = element_blank(),
    axis.text = element_blank(),
    axis.ticks = element_blank(),
    legend.box = "vertical",
    legend.box.spacing = unit(0, "cm"),
    legend.spacing.y = unit(0, "cm"),
    legend.margin = margin(t = 0, b = 0),
    legend.title = element_text(size = 10),
    legend.text = element_text(size = 9),
    legend.key.height = unit(0.4, "cm")
  ) +
  guides(
    linetype = guide_legend(order = 1, override.aes = list(color = "yellow", linewidth = 1)),
    fill = guide_legend(order = 2, override.aes = list(alpha = 0.7, shape = 22, color = NA, size = 5)),
    shape = guide_legend(order = 3, override.aes = list(size = 4, stroke = 1.2)),
    size = guide_legend(order = 4, override.aes = list(shape = 16, stroke = 0.8))
  )

# CA inset
map_centroid_coords <- map_site_sf %>%
  st_coordinates() %>%
  as.data.frame() %>%
  summarise(longitude = mean(X), latitude = mean(Y))

map_CA <- map("state", plot = FALSE, fill = TRUE) %>%
  st_as_sf() %>%
  filter(ID == "california")

map_overview <- ggplot() +
  geom_sf(data = map_CA, fill = "gray90", color = "gray60", size = 0.3) +
  geom_point(data = map_centroid_coords, aes(x = longitude, y = latitude),
             color = "red", size = 3, shape = 15) +
  theme_void() +
  theme(
    panel.border = element_rect(color = "black", fill = NA, size = 0.5),
    plot.background = element_rect(fill = "white", color = NA)
  )

map_final <- ggdraw() +
  draw_plot(map_main) +
  draw_plot(map_overview, x = 0.68, y = 0.647, width = 0.3, height = 0.3)

map_final

ggsave(here("figures", "deployment_map_pt_2_v2.png"), plot = map_final, width = 10, height = 6, dpi = 300, bg = "white")
```


# Part 1: Winter Study

## Load Data
```{r}
##load study part 1 data from February 2021
clean_pt_1_obs <- read_excel(here("data","originalMasterHibernaculaAnalysis.xlsx"), sheet = "Master.Data.Cleaned")

# Recode raw Excel value to new terminology
clean_pt_1_obs$feature_type_methodology[clean_pt_1_obs$feature_type_methodology == "Constructed Hibernacula"] <- "Artificial Refuge"

#load AR site coordinates
hibernacula_coords <- read_csv(here("data", "hibernacula_coordinates.csv")) |> 
  mutate(placename = str_to_upper(placename))
```

## Clean and Prepare Data
```{r}
#join AR site coordinates to clean_pt_1_obs
clean_pt_1_obs <- clean_pt_1_obs |> 
  left_join(hibernacula_coords, by = c("placename"))|> 
  filter(!is.na(common_name)) |> 
  mutate(
    obs_start_time = hms::as_hms(obs_start_time),
    common_name = common_name |>
      str_replace_all("Racoon", "Northern Raccoon") |>
      str_replace_all("California Towee", "California Towhee") |>
      str_replace_all("Says Phoebe", "Say's Phoebe") |>
      str_replace_all("Coopers Hawk", "Cooper's Hawk") |>
      str_replace_all("Hermit thrush", "Hermit Thrush") |>
      str_replace_all("Spotted Skunk", "Western Spotted Skunk") |>
      str_replace_all("Western Skink", "Western skink") |>
      str_replace_all("Mouse", "North American Deermouse") |>
      str_replace_all("Western Fence Lizard", "Western fence lizard") |> 
      str_replace_all("American Pipit", "Buff-bellied Pipit")
  ) |> 
  left_join(site_to_habitat_crosswalk, by = c("placename" = "site")) |> 
  mutate(
    deployment_duration = 5, ## ASSUMES ALL FEBRUARY DEPLOYMENTS WERE 5 DAYS LONG
    study_part = 1 ## tags all of these sequences as part of the first part of the study
  ) |> 
  group_by(placename,camera_name) |> 
  mutate(
    start_date = min(obs_start_date, na.rm = TRUE),
    end_date = max(obs_start_date, na.rm = TRUE),
    calculated_deployment_duration = as.numeric(end_date - start_date) ### this shows that some sites may not have had 5 full days of trapping: H12, H13, H33, H48, H64 (4 days each); H25 (3 days), H21 2 (days)
  ) |> 
  ungroup() |> 
  mutate( ## manual corrections for start/end dates based on review of first/last recorded images
    start_date = case_when(
      placename == "H21" & camera_name == "C12" & study_part == 1 ~ as.Date("2021-02-18"),
      placename == "H25" & camera_name == "C2" & study_part == 1 ~ as.Date("2021-02-04"),
      placename == "H48" & camera_name == "C10" & study_part == 1 ~ as.Date("2021-02-11"),
      placename == "H11" & camera_name == "C11" & study_part == 1 ~ as.Date("2021-02-04"),
      placename == "H11" & camera_name == "C12" & study_part == 1 ~ as.Date("2021-02-04"),
      placename == "H58" & camera_name == "C09" & study_part == 1 ~ as.Date("2021-02-04"),
      placename == "H19" & camera_name == "C05" & study_part == 1 ~ as.Date("2021-02-18"),
      placename == "H34" & camera_name == "C05" & study_part == 1 ~ as.Date("2021-02-04"),
      TRUE ~ start_date
    ),
    end_date = case_when(
      placename == "H21" & camera_name == "C12" & study_part == 1 ~ as.Date("2021-02-23"),
      placename == "H25" & camera_name == "C2" & study_part == 1 ~ as.Date("2021-02-09"),
      placename == "H12" & camera_name == "C11" & study_part == 1 ~ as.Date("2021-03-09"),
      placename == "H12" & camera_name == "C12" & study_part == 1 ~ as.Date("2021-03-09"),
      placename == "H13" & camera_name == "C2" & study_part == 1 ~ as.Date("2021-02-16"),
      placename == "H16" & camera_name == "C9" & study_part == 1 ~ as.Date("2021-03-09"),
      placename == "H11" & camera_name == "C12" & study_part == 1 ~ as.Date("2021-02-09"),
      placename == "H58" & camera_name == "C09" & study_part == 1 ~ as.Date("2021-02-09"),
      placename == "H47" & camera_name == "C08" & study_part == 1 ~ as.Date("2021-03-02"),
      placename == "H59" & camera_name == "C04" & study_part == 1 ~ as.Date("2021-02-16"),
      TRUE ~ end_date
    )
  ) |> 
   group_by(placename,camera_name) |> 
  mutate(
    calculated_deployment_duration = as.numeric(end_date - start_date),
    deployment_duration = calculated_deployment_duration) |> 
  ungroup() |> 
filter(!(placename == "H41" & camera_name == "C6"), ## removing deployment b/c camera was set at bad angle)
       !(placename == "H23" & camera_name == "C3"),  ## emoving deployment b/c camera was set at bad angle)
       !(placename == "H24" & camera_name == "C1")) |>  ## removing  deployment because it failed on day 1
  dplyr::select(-calculated_deployment_duration)

clean_pt_1_obs <- clean_pt_1_obs |> 
  mutate(row_id = row_number())
```


```{r}
## Remove Duplicate Observations
  clean_pt_1_obs <- clean_pt_1_obs |> 
    mutate(
      row_id = row_number(),
      start_time = as.POSIXct(obs_start_date) + as.numeric(obs_start_time),
      truncated_time = floor_date(start_time, unit = "minute"),
      end_time = start_time + seconds(60)  # assume we want to remove duplicates within the minute
    )

  # Self-join within group to find same-minute, same-species, different-camera overlaps
  overlap_pairs_part_1 <- clean_pt_1_obs |> 
    inner_join(clean_pt_1_obs, by = c("placename", "obs_start_date", "truncated_time", "common_name")) |> 
    filter(
      camera_name.x != camera_name.y,
      row_id.x != row_id.y
    ) |> 
    distinct(row_id.x, row_id.y, .keep_all = TRUE)
  
  rows_to_remove_part_1 <- overlap_pairs_part_1 |> 
    mutate(
      keep_x = if_else(group_size.x > group_size.y, TRUE,
                if_else(group_size.x < group_size.y, FALSE,
                  end_time.x > end_time.y))  # later end_time wins if tie
    ) |> 
    filter(!keep_x) |> 
    pull(row_id.x) |> 
    unique()
  
  clean_pt_1_obs <- clean_pt_1_obs |> 
    filter(!row_id %in% rows_to_remove_part_1)

## Deployment Duration
placename_summary_pt_1 <- clean_pt_1_obs |> 
  filter(!str_detect(camera_name, "&|,")) |> 
  group_by(
    placename, camera_name, start_date, end_date, deployment_duration,
    feature_type_methodology, habitat_type, notable_entrances, trail, study_part,latitude,longitude
  ) |> 
  summarize(.groups = "drop")

##SUMMARIZE FEBRUARY OBSERVATIONS BY SPECIES PER HOUR PER PLACE
hourly_presence_pt_1 <- clean_pt_1_obs |>
  # Create hour-block and date
  mutate(
    obs_hour = hour(obs_start_time),
    obs_day = as.Date(obs_start_date)
  ) |> 
  #Count unique species per hour block, day, and place
  group_by(placename, obs_day, obs_hour, feature_type_methodology,common_name,habitat_type,notable_entrances,trail,study_part,start_date,end_date,deployment_duration) |> 
  summarize(n = 1, .groups = "drop")  # Forces 1 row per unique species in each hour/day block
```

# Combined Analysis (Winter + Spring) 
```{r}
## Join Effort and Detections
deployments_summary_pt_1_and_2 <- placename_summary_pt_1 |> 
  full_join(clean_pt_2_dep, by = c("placename", "camera_name","start_date", "end_date", "deployment_duration", "feature_type_methodology", "habitat_type","notable_entrances", "trail", "study_part","latitude","longitude"))

write.csv(deployments_summary_pt_1_and_2, here("data","deployments_summary_pt_1_and_2.csv"), row.names = FALSE)


### Hourly Presence Reduction
hourly_presence_pt_2 <- clean_pt_2_seq_dep |>
  mutate(
    obs_hour = hour(obs_start_time),
    obs_day = as.Date(obs_start_date)
  ) |>
  group_by(
    placename,
    obs_day,
    obs_hour,
    feature_type_methodology,
    common_name,
    habitat_type,
    notable_entrances,
    trail,
    study_part,
    start_date,
    end_date,
    deployment_duration
  ) |>
  summarize(n = 1, .groups = "drop")

## Combine Hourly Presence (Winter + Spring)
hourly_presence_pt_1_and_2 <- bind_rows(hourly_presence_pt_1, hourly_presence_pt_2)

# Count unique species per hour, feature, and study period
hourly_presence_pt_1_and_2_summary <- hourly_presence_pt_1_and_2 |>
  group_by(study_part, feature_type_methodology, obs_hour) |>
  summarise(species_detections = n(), .groups = "drop")

feature_type_effort_summary_pt_1 <- placename_summary_pt_1 %>%
    group_by(feature_type_methodology) %>%
    summarize(
        total_camera_days = sum(deployment_duration, na.rm = TRUE),
        total_camera_hours = total_camera_days * 24,
        .groups = "drop"
    )
```

## Data Visualizations:
### Species Residency Status by Feature Type

```{r fig-residency-status-by-feature}
#| fig-label: fig-residency-status-by-feature
#| fig-cap: "**Species residency status by feature type.** Detection rates (avg. # of hours each species is detected per camera day) for resident and transient species across different feature types. Points represent individual species. Species with detection rates above 0.25 detections per camera day are classified as residents. Resident species are colored, Black diamonds show mean detection rates with 95% confidence intervals."

# 1. Camera effort summed across study parts

feature_type_effort_summary_pt_1_and_2 <- bind_rows(
    feature_type_effort_summary_pt_1 %>% mutate(study_part = 1),
    feature_type_effort_summary_pt_2 %>% mutate(study_part = 2)) |> 
  group_by(feature_type_methodology)

feature_type_effort_combined_summary_pt_1_and_2 <- feature_type_effort_summary_pt_1_and_2 %>%
  group_by(feature_type_methodology) %>%
  summarize(total_camera_days = sum(total_camera_days, na.rm = TRUE),
 total_camera_hours = sum(total_camera_hours, na.rm = TRUE), .groups = "drop")

# 2. Calculate detections per camera hour for each species-feature combo
species_hourly_presence_by_feature_pt_1_and_2 <- hourly_presence_pt_1_and_2 %>%
  group_by(feature_type_methodology, common_name) %>%
  summarise(detections = n(), .groups = "drop") %>%
  left_join(feature_type_effort_combined_summary_pt_1_and_2, by = "feature_type_methodology") %>%
  mutate(detections_per_camera_day = detections / total_camera_days)

# 3. Assign residency status
residency_threshold <- 0.25
species_hourly_presence_by_feature_pt_1_and_2 <- species_hourly_presence_by_feature_pt_1_and_2 %>%
  mutate(residency_status = ifelse(detections_per_camera_day > residency_threshold, "resident", "transient"))


# 5. Calculate summary statistics (mean, CI) for each feature type and residency status
summary_stats_species_hourly_presence_by_feature_pt_1_and_2 <- species_hourly_presence_by_feature_pt_1_and_2 %>%
  group_by(feature_type_methodology, residency_status) %>%
  summarise(
    mean = mean(detections_per_camera_day, na.rm = TRUE),
    se = sd(detections_per_camera_day, na.rm = TRUE) / sqrt(n()),
    n = n(),
    lower = mean - qt(0.975, df = n - 1) * se,
    upper = mean + qt(0.975, df = n - 1) * se,
    .groups = "drop"
  )

max_y <- max(species_hourly_presence_by_feature_pt_1_and_2$detections_per_camera_day, na.rm = TRUE)
max_y <- ceiling(max_y * 1.3 * 10) / 10  # 20% buffer and round to nearest 0.1

# 6. Create a plot for each feature type
feature_types <- c("Artificial Refuge", "Boulder", "Log")

# Step 1: Assign "Other" label to species not in the palette
species_hourly_presence_by_feature_pt_1_and_2 <- species_hourly_presence_by_feature_pt_1_and_2 %>%
  mutate(
    species_fill = ifelse(common_name %in% names(species_palette_pt_2),
                          common_name,
                          "Other")
  )

# Step 2: Reorder species so "Other" comes last
species_levels <- c(sort(setdiff(unique(species_hourly_presence_by_feature_pt_1_and_2$species_fill), "Other")), "Other")
species_hourly_presence_by_feature_pt_1_and_2 <- species_hourly_presence_by_feature_pt_1_and_2 %>%
  mutate(species_fill = factor(species_fill, levels = species_levels))

# Step 2: Add 'Other' color to palette
extended_palette <- species_palette_pt_2
if (!"Other" %in% names(extended_palette)) {
  extended_palette["Other"] <- "#a6cee3"
}

extended_palette <- extended_palette[species_levels]

plots <- lapply(seq_along(feature_types), function(i) {
  ft <- feature_types[i]

  p <- ggplot(
    species_hourly_presence_by_feature_pt_1_and_2 %>% filter(feature_type_methodology == ft),
    aes(x = residency_status, y = detections_per_camera_day)
  ) +
    geom_jitter(aes(fill = species_fill, shape = species_fill), color = "black", width = 0.25, alpha = .85, size = 5, show.legend = TRUE) +
    geom_errorbar(
      data = summary_stats_species_hourly_presence_by_feature_pt_1_and_2 %>% filter(feature_type_methodology == ft),
      aes(y = mean, ymin = lower, ymax = upper),
      width = 0.5, color = "black", linewidth = 0.7
    ) +
    geom_point(
      data = summary_stats_species_hourly_presence_by_feature_pt_1_and_2 %>% filter(feature_type_methodology == ft),
      aes(y = mean),
      shape = 9, size = 5,
      fill = "white", color = "black", stroke = 1.2
    ) +
    scale_fill_manual(values = extended_palette, name="Species") +
    scale_shape_manual(values = c(21, 22, 23, 24, 25), name="Species") +
    scale_y_continuous(limits = c(0, max_y)) +
    labs(
      title = paste(ft)
    ) +
    theme_minimal() +
    theme(
      axis.text.x = element_text(vjust = 1),
      plot.title = element_text(hjust = 0.5, size = 12, face = "bold")  # Center and style the title
    )
  
  # Only add y-axis label to the first (leftmost) plot
  if(i == 1) {
    p <- p + labs(y = "Avg. # of Hours Detected per Camera Day")
  } else {
    p <- p + labs(y = "")
  }
  
  # Only add x-axis label to the middle chart
  middle_index <- ceiling(length(feature_types) / 2)
  if(i == middle_index) {
    p <- p + labs(x = "Residency Status")
  } else {
    p <- p + labs(x = "")
  }
  
  return(p)
})

# 7. Display all plots side by side
p4 <- wrap_plots(plots, guides = "collect") &
  theme_minimal(base_size = 16) &  # bigger text everywhere
  theme(
    axis.text.x = element_text(size = 18, angle = 0, vjust = 0.5),
    axis.text.y = element_text(size = 16),
    axis.title.x = element_text(size = 18, face = "bold"),
    axis.title.y = element_text(size = 18, face = "bold"),
    plot.title = element_text(size = 20, face = "bold", hjust = 0.5),
    legend.position = "bottom",
    legend.box = "horizontal",
    legend.title = element_text(size = 16),
    legend.text = element_text(size = 14)
  ) &
  guides(fill = guide_legend(nrow = 1, byrow = TRUE))
p4
ggsave(here("figures","species_residency_status_by_feature_type_pt_1_and_2.png"), 
       plot = p4, 
       width = 12, height = 6, dpi = 300, bg = "white")
```

## Combined Deployment Map
```{r}
# Extract the zip file 
unzip(here("data", "ncos_shp.zip"), exdir = here("data")) 
# List the extracted files to see what's available 
list.files(here("data"), pattern = "\\.shp$") 
# Read the shapefile (replace "filename" with the actual .shp filename) 
open_space_boundary <- st_read(here("data", "ncos_shp.shp"))
# Transform boundary to WGS84 to match your study area
open_space_boundary <- st_transform(open_space_boundary, crs = 4326)

# Create combined dataset from both study parts
placenames_for_deployment_map_step1 <- bind_rows(
    placename_summary_pt_1 %>%
      group_by(placename, latitude, longitude, feature_type_methodology, habitat_type) %>%
      summarise(study_part = first(study_part), .groups = "drop"),
    placename_summary_pt_2 %>%
      group_by(placename, latitude, longitude, feature_type_methodology, habitat_type) %>%
      summarise(study_part = first(study_part), .groups = "drop")
  )


# Combined Map for Both Parts of the Study
placenames_for_deployment_map <- placenames_for_deployment_map_step1 %>%
  mutate(
    latitude = round(latitude, 4),
    longitude = round(longitude, 4)
  ) %>%
  group_by(placename, latitude, longitude, feature_type_methodology) %>%
  summarise(
    study_part_combined = if_else(n() > 1, "both", as.character(min(study_part))),
    .groups = "drop"
  )


# Convert site_summary to sf object
map_site_sf <- st_as_sf(placenames_for_deployment_map, coords = c("longitude", "latitude"), crs = 4326)

# Get bbox from shapefile and expand
bbox_expanded <- st_bbox(open_space_boundary)
buffer_val <- 0  # tweak this until boundary fits comfortably
bbox_expanded <- bbox_expanded + c(-buffer_val, -buffer_val, buffer_val, buffer_val)

# Convert to sf polygon for tile download
bbox_sf <- st_as_sfc(bbox_expanded)

# Download satellite tiles using expanded bbox
map_basemap <- get_tiles(bbox_sf, provider = "Esri.WorldImagery", zoom = 18)

# Load and process habitat layer
ncos_habitats <- st_read(here("data", "ncos_habitats_june_2021", "NCOS_Habitats_June2021_Simple.shp"))
ncos_habitats <- st_make_valid(ncos_habitats)
ncos_habitats <- st_transform(ncos_habitats, crs = 4326)

# Simplify habitat categories and dissolve by group
ncos_habitats <- ncos_habitats |>
  mutate(habitat_simple = case_when(
    str_detect(Habitat, "Grassland") ~ "Grassland",
    Habitat == "Upland Mosaic" ~ "Grassland",
    str_detect(Habitat, "Marsh|marsh") ~ "Marsh",
    str_detect(Habitat, "Scrub") ~ "Scrub",
    Habitat == "Oak Woodland/Chaparral" ~ "Scrub",
    TRUE ~ "Other"
  )) |>
  group_by(habitat_simple) |>
  summarise(geometry = st_union(geometry), .groups = "drop")

# Clip habitat layer to NCOS boundary
ncos_habitats <- st_intersection(ncos_habitats, open_space_boundary)

# Filter out "Other" habitat type
ncos_habitats_filtered <- ncos_habitats |> filter(habitat_simple != "Other")

# Define simplified habitat colors
habitat_colors <- c(
  "Grassland" = "#9ACD32",
  "Marsh" = "#2E8B57",
  "Scrub" = "#9370DB"
)

# Convert raster to data.frame for ggplot
map_rgb_df <- as.data.frame(map_basemap, xy = TRUE)
colnames(map_rgb_df) <- c("x", "y", "red", "green", "blue")
map_rgb_df$hex <- rgb(map_rgb_df$red, map_rgb_df$green, map_rgb_df$blue, maxColorValue = 255)

  # Create combined shape variable: feature type + study participation
  # Reviewer: squares for AR (open=one study, filled=both), triangles for boulders, circles for logs
  map_site_sf <- map_site_sf %>%
    mutate(
      point_label = case_when(
        feature_type_methodology == "Artificial Refuge" & study_part_combined == "both" ~ "Refuge (both studies)",
        feature_type_methodology == "Artificial Refuge" ~ "Refuge (one study)",
        feature_type_methodology == "Boulder" ~ "Boulder",
        feature_type_methodology == "Log" ~ "Log",
        TRUE ~ as.character(feature_type_methodology)
      ),
      point_label = factor(point_label, levels = c(
        "Refuge (both studies)", "Refuge (one study)", "Boulder", "Log"
      ))
    )

  # Shape encoding: filled square=15, open square=0, triangle=2, circle=1
  point_shapes <- c(
    "Refuge (both studies)" = 15,
    "Refuge (one study)" = 0,
    "Boulder" = 2,
    "Log" = 1
  )

  # Main satellite deployment map
  map_main <- ggplot() +
    geom_raster(data = map_rgb_df, aes(x = x, y = y, fill = hex)) +
    scale_fill_identity() +

    # Habitat polygons
    ggnewscale::new_scale_fill() +
    geom_sf(data = ncos_habitats_filtered, aes(fill = habitat_simple), color = NA, alpha = 0.5) +
    scale_fill_manual(values = habitat_colors, name = "Habitat Type",
                      guide = guide_legend(order = 2)) +

    # NCOS boundary
    geom_sf(data = open_space_boundary, aes(linetype = "NCOS Boundary"),
            color = "yellow", fill = NA, linewidth = 2) +

    # Deployment points — shape only, no fill color
    geom_sf(data = map_site_sf, aes(shape = point_label),
            color = "black", size = 4, stroke = 1.2) +

    scale_shape_manual(
      name = "Feature Type",
      values = point_shapes
    ) +
    scale_linetype_manual(
      name = "",
      values = c("NCOS Boundary" = "solid")
    ) +

    coord_sf(expand = TRUE) +
    annotation_scale(location = "bl", pad_x = unit(0.45, "in"), pad_y = unit(0.3, "in"),
                     width_hint = 0.25, text_col = "white") +
    annotation_north_arrow(location = "tr", which_north = "true",
                           pad_x = unit(0.40, "in"), pad_y = unit(0.35, "in"),
                           style = north_arrow_fancy_orienteering(
                             fill = c("white", "white"),
                             line_col = "white",
                             text_col = "white"
                           )) +

    theme_minimal(base_size = 12) +
    theme(
      panel.grid = element_blank(),
      legend.position = "bottom",
      legend.box = "horizontal",
      legend.box.just = "center",
      legend.box.margin = margin(t = -20, r = 0, b = 0, l = 0),
      plot.margin = margin(t = 5, r = 5, b = 5, l = 5),
      axis.title = element_blank(),
      axis.text = element_blank(),
      axis.ticks = element_blank(),
      legend.title = element_text(size = 14, face = "bold"),
      legend.text = element_text(size = 12),
      legend.spacing.x = unit(0.5, "cm"),
      legend.direction = "vertical"
    ) +

    guides(
      shape = guide_legend(order = 1, override.aes = list(size = 4, stroke = 1.2)),
      fill = guide_legend(order = 2, override.aes = list(alpha = 0.7, shape = 22, color = NA, size = 5)),
      linetype = guide_legend(order = 3, override.aes = list(color = "yellow", linewidth = 2))
    )

# Calculate study area centroid
map_centroid_coords <- map_site_sf %>%
  st_coordinates() %>%
  as.data.frame() %>%
  summarise(longitude = mean(X), latitude = mean(Y))

# Overview map of CA with study area location
map_CA <- map("state", plot = FALSE, fill = TRUE) %>%
  st_as_sf() %>%
  filter(ID == "california")

map_overview <- ggplot() +
  geom_sf(data = map_CA, fill = "gray90", color = "gray60", size = 0.3) +
  geom_point(data = map_centroid_coords, aes(x = longitude, y = latitude), 
             color = "red", size = 3, shape = 15) +
  theme_void() +
  theme(
    panel.border = element_rect(color = "black", fill = NA, size = 0.5),
    plot.background = element_rect(fill = "white", color = NA)
  )

# Combine maps with proper positioning
map_final <- ggdraw() +
  draw_plot(map_main) +
  draw_plot(map_overview, x = 0.69, y = .25 , width = 0.3, height = 0.3)

map_final

ggsave(here("figures","deployment_map_combined.png"), plot = map_final, width = 10, height = 6, dpi = 300, bg = "white")
```

### Map: Two-Panel (Winter / Spring)
```{r fig-deployment-map-2panel}
#| fig-cap: "**Camera trap deployment locations during winter (left) and spring (right) studies.** Point shape indicates feature type: square = artificial refuge, triangle = boulder, circle = log. Habitat types shown as semi-transparent overlays. Inset shows study location within California."
#| fig-width: 14
#| fig-height: 7

# Split sites by study period (use step1 data which has separate rows per study_part)
sites_winter <- placenames_for_deployment_map_step1 %>%
  filter(study_part == 1) %>%
  st_as_sf(coords = c("longitude", "latitude"), crs = 4326)

sites_spring <- placenames_for_deployment_map_step1 %>%
  filter(study_part == 2) %>%
  st_as_sf(coords = c("longitude", "latitude"), crs = 4326)

# Shape encoding
point_shapes_map <- c("Artificial Refuge" = 15, "Boulder" = 2, "Log" = 1)

# Shared map-building function
build_panel <- function(site_data, panel_title) {
  ggplot() +
    geom_raster(data = map_rgb_df, aes(x = x, y = y, fill = hex)) +
    scale_fill_identity() +
    ggnewscale::new_scale_fill() +
    geom_sf(data = ncos_habitats_filtered, aes(fill = habitat_simple), color = NA, alpha = 0.5) +
    scale_fill_manual(values = habitat_colors, name = "Habitat Type") +
    geom_sf(data = open_space_boundary, color = "yellow", fill = NA, linewidth = 1.5) +
    geom_sf(data = site_data, aes(shape = feature_type_methodology), color = "black", size = 3.5, stroke = 1.2) +
    scale_shape_manual(values = point_shapes_map, name = "Feature Type") +
    coord_sf(expand = TRUE) +
    labs(title = panel_title) +
    annotation_scale(location = "bl", pad_x = unit(0.3, "in"), pad_y = unit(0.2, "in"),
                     width_hint = 0.2, text_col = "white") +
    theme_minimal(base_size = 11) +
    theme(
      panel.grid = element_blank(),
      axis.title = element_blank(),
      axis.text = element_blank(),
      axis.ticks = element_blank(),
      plot.title = element_text(size = 14, face = "bold", hjust = 0.5),
      legend.position = "none"
    )
}

panel_winter <- build_panel(sites_winter, "Winter (Feb–Mar 2021)")
panel_spring <- build_panel(sites_spring, "Spring (Apr–May 2021)")

# Shared legend from spring panel (has all 3 feature types)
legend_plot <- ggplot() +
  geom_raster(data = map_rgb_df[1, ], aes(x = x, y = y, fill = hex)) +
  scale_fill_identity() +
  ggnewscale::new_scale_fill() +
  geom_sf(data = ncos_habitats_filtered, aes(fill = habitat_simple), color = NA, alpha = 0.5) +
  scale_fill_manual(values = habitat_colors, name = "Habitat Type") +
  geom_sf(data = sites_spring, aes(shape = feature_type_methodology), color = "black", size = 3.5, stroke = 1.2) +
  scale_shape_manual(values = point_shapes_map, name = "Feature Type") +
  theme(legend.position = "bottom", legend.box = "horizontal",
        legend.title = element_text(size = 11, face = "bold"),
        legend.text = element_text(size = 10)) +
  guides(
    fill = guide_legend(order = 1, override.aes = list(alpha = 0.7, shape = 22, color = NA, size = 5)),
    shape = guide_legend(order = 2, override.aes = list(size = 4, stroke = 1.2))
  )

shared_legend <- cowplot::get_legend(legend_plot)

# CA inset
map_overview_small <- ggplot() +
  geom_sf(data = map_CA, fill = "gray90", color = "gray60", size = 0.3) +
  geom_point(data = map_centroid_coords, aes(x = longitude, y = latitude),
             color = "red", size = 2.5, shape = 15) +
  theme_void() +
  theme(
    panel.border = element_rect(color = "black", fill = NA, size = 0.5),
    plot.background = element_rect(fill = "white", color = NA)
  )

# Assemble: two panels side by side, shared legend below, CA inset overlaid
two_panels <- plot_grid(panel_winter, panel_spring, ncol = 2, rel_widths = c(1, 1))
map_2panel <- plot_grid(two_panels, shared_legend, ncol = 1, rel_heights = c(1, 0.08))

map_2panel_final <- ggdraw(map_2panel) +
  draw_plot(map_overview_small, x = 0.82, y = 0.75, width = 0.16, height = 0.22)

map_2panel_final

ggsave(here("figures", "deployment_map_2panel.png"), plot = map_2panel_final,
       width = 14, height = 7, dpi = 300, bg = "white")
```

## Entrances by Feature Type

```{r fig-entrances-by-feature}
#| fig-label: fig-entrances-by-feature
#| fig-cap: "**Number of entrances by feature type.** Boxplots show the distribution of the number of entrances per feature type. Points represent individual features. The horizontal line indicates the median number of entrances."

placename_summary_pt_1_and_2 <- deployments_summary_pt_1_and_2 %>%
  filter(!is.na(notable_entrances)) %>%
  group_by(placename, feature_type_methodology,latitude,longitude) %>%
  summarise(
    notable_entrances = max(notable_entrances),
    .groups = "drop"
  )

# Step 1: Summarize the table
entrances_by_feature_type <- placename_summary_pt_1_and_2 %>%
  group_by(feature_type_methodology) %>%
  summarise(
    mean_entrances = mean(notable_entrances, na.rm = TRUE),
    median_entrances = median(notable_entrances, na.rm = TRUE),
    sd_entrances = sd(notable_entrances, na.rm = TRUE),
    n = n(),
    .groups = "drop"
  )

# Step 2: Define output paths
html_file <- here("misc", "entrances_by_feature_type_table.html")
png_file  <- here("figures", "entrances_by_feature_type_table_pt_1_and_2.png")

# Step 3: Create and save HTML kable
kable(entrances_by_feature_type, format = "html", digits = 2, caption = "Entrances by Feature Type") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = F) %>%
  save_kable(file = html_file)

# Step 4: Save as PNG using webshot
tryCatch(
  webshot(html_file, file = png_file, zoom = 2, vwidth = 800, vheight = 600),
  error = function(e) message("Webshot skipped: ", e$message)
)

# Create the boxplot
p5 <- ggplot(placename_summary_pt_1_and_2, aes(x = feature_type_methodology, y = notable_entrances)) +
  geom_boxplot(outlier.shape = NA, fill = "lightgray", alpha = 0.7) +
  geom_jitter(aes(color = feature_type_methodology), size = 2, alpha = 0.6, width = 0.2) +
  # Add dashed median lines for each feature type
  geom_segment(
    data = entrances_by_feature_type,
    aes(
      x = feature_type_methodology, xend = feature_type_methodology,
      y = median_entrances, yend = median_entrances
    ),
    inherit.aes = FALSE,
    linetype = "dashed", color = "red", linewidth = 1
  ) +
  labs(
    title = "Number of Entrances by Feature Type",
    x = "Feature Type",
    y = "Number of Entrances"
  ) +
  scale_color_manual(values = c("Artificial Refuge" = "#fc8d62", "Boulder" = "#66c2a5", "Log" = "#8da0cb")) +
  theme_minimal(base_size = 14) +
  theme(
    legend.position = "none",)

p5

ggsave(here("figures","entrances_by_feature_type_pt_1_and_2.png"), 
       plot = p5, 
       width = 8, height = 6, dpi = 300, bg = "white")
```

```{r}
# Statistical test: Pairwise Wilcoxon test for number of entrances by feature type
pairwise.wilcox.test(
    placename_summary_pt_1_and_2$notable_entrances,
    placename_summary_pt_1_and_2$feature_type_methodology,
    p.adjust.method = "holm"
)
```
    
## Species List (Winter)

```{r tbl-species-list-pt1}
#| tbl-cap: "Species List for Part 1 of the Study. Only includes wildlife identified to species level. Table is sorted by number of detections." 

# Calculate total unique features (placename) per feature_type
total_features_pt_1 <- clean_pt_1_obs %>%
  distinct(feature_type_methodology, placename) %>%
  group_by(feature_type_methodology) %>%
  summarise(total_features_pt_1 = n(), .groups = "drop")

# Summarize sum(group_size) and count unique visited placename per species per feature type
species_summary_pt_1 <- clean_pt_1_obs %>%
  group_by(common_name, feature_type_methodology) %>%
  summarise(
    count_observed = sum(group_size, na.rm = TRUE),
    features_visited = n_distinct(placename),
    .groups = "drop"
  )

# Pivot counts wide
species_counts_wide_pt_1 <- species_summary_pt_1 %>%
  dplyr::select(-features_visited) %>%
  pivot_wider(
    names_from = feature_type_methodology,
    values_from = count_observed,
    values_fill = 0,
    names_prefix = "count_observed_at_"
  ) |> 
  clean_names()

# Pivot features visited wide
species_features_wide_pt_1 <- species_summary_pt_1 %>%
  dplyr::select(common_name, feature_type_methodology, features_visited) %>%
  pivot_wider(
    names_from = feature_type_methodology,
    values_from = features_visited,
    values_fill = 0,
    names_prefix = "features_visited_at_"
  ) |> 
  clean_names()

# Combine counts and features visited
species_combined_pt_1 <- species_counts_wide_pt_1 %>%
  full_join(species_features_wide_pt_1, by = c("common_name"))

# Extract total features per feature type as named vector
total_features_pt_1_vec <- total_features_pt_1 %>%
  deframe()  # named vector: names are feature_type_methodology, values total_features_pt_1

# Calculate percent visited by species at each feature type
species_final_pt_1 <- species_combined_pt_1 %>%
  mutate(
    percent_hibernacula_visited = 100 * features_visited_at_artificial_refuge / total_features_pt_1_vec["Artificial Refuge"]
  )

camera_days_vec_pt_1 <- feature_type_effort_summary_pt_1 %>%
  dplyr::select(feature_type_methodology, total_camera_days) %>%
  pull(total_camera_days) %>%
  setNames(feature_type_effort_summary_pt_1$feature_type_methodology)

total_camera_days_pt_1 <- sum(camera_days_vec_pt_1)

species_final_pt_1 <- species_final_pt_1 %>%
  mutate(
    rate_per_cam_day_hibernacula = count_observed_at_artificial_refuge / camera_days_vec_pt_1["Artificial Refuge"]
  )

# Create formatted table
t1_pt1 = species_final_pt_1 %>%
  dplyr::select(
  #  'Class' = class,
    'Common Name' = common_name,
  #  'Scientific Name' = genus_species,
    'Detections' = count_observed_at_artificial_refuge,
    '% Visited' = percent_hibernacula_visited,
    'Per Cam.-Day' = rate_per_cam_day_hibernacula,
  ) %>%
  arrange(desc(`Detections`), `Common Name`) %>%
  kable(
    caption = "Species Observations and Percent Feature Visited by Feature Type (Part 1)",
    align = c("l", "l", "l", rep("c", 12)),
    digits = 3
  ) %>%
  kable_styling(
    full_width = FALSE,
    position = "left",
    bootstrap_options = c("striped", "hover", "condensed")
  ) %>%
  add_header_above(c(" " = 1, "Artificial Refuges" = 3))

tryCatch({
  save_kable(t1_pt1, here("misc","species_table_pt1.html"))
  webshot(here("misc","species_table_pt1.html"), here("figures","species_table_pt1.png"), zoom = 2, vwidth = 550)
}, error = function(e) message("Table save/screenshot skipped: ", e$message))

```
    
## Species List: Hourly Presence (Combined)
```{r}
#| label: tbl-species-list-hourly_pt_1_and_2
#| tbl-cap: "Species List from Hourly Presence Data (Parts 1 and 2 Combined). Includes species-level wildlife identifications, sorted by class then total detections." 

# Calculate total unique features (placename) per feature_type
total_features_hourly <- hourly_presence_pt_1_and_2 %>%
  distinct(feature_type_methodology, placename) %>%
  group_by(feature_type_methodology) %>%
  summarise(total_features_hourly = n(), .groups = "drop")

# Summarize total detections and number of unique features visited
species_summary_hourly <- hourly_presence_pt_1_and_2 %>%
  group_by(common_name,feature_type_methodology) %>%
  summarise(
    count_observed = sum(n, na.rm = TRUE),
    features_visited = n_distinct(placename),
    .groups = "drop"
  )

# Pivot detection counts wide
species_counts_wide_hourly <- species_summary_hourly %>%
  dplyr::select(-features_visited) %>%
  pivot_wider(
    names_from = feature_type_methodology,
    values_from = count_observed,
    values_fill = 0,
    names_prefix = "count_observed_at_"
  ) %>%
  clean_names()

# Pivot features visited wide
species_features_wide_hourly <- species_summary_hourly %>%
  dplyr::select(common_name, feature_type_methodology, features_visited) %>%
  pivot_wider(
    names_from = feature_type_methodology,
    values_from = features_visited,
    values_fill = 0,
    names_prefix = "features_visited_at_"
  ) %>%
  clean_names()

# Combine counts and features visited
species_combined_hourly <- species_counts_wide_hourly %>%
  full_join(species_features_wide_hourly, by = c("common_name"))

# Named vector of total features per feature type
total_features_hourly_vec <- total_features_hourly %>%
  deframe()

# Calculate percent visited for each feature type
species_final_hourly <- species_combined_hourly %>%
  mutate(
    percent_hibernacula_visited = 100 * features_visited_at_artificial_refuge / total_features_hourly_vec["Artificial Refuge"],
    percent_boulders_visited = 100 * features_visited_at_boulder / total_features_hourly_vec["Boulder"],
    percent_logs_visited = 100 * features_visited_at_log / total_features_hourly_vec["Log"]
  )

# Camera effort vector for denominator in rates
camera_days_vec_hourly <- feature_type_effort_combined_summary_pt_1_and_2 %>%
  dplyr::select(feature_type_methodology, total_camera_days) %>%
  pull(total_camera_days) %>%
  setNames(feature_type_effort_combined_summary_pt_1_and_2 $feature_type_methodology)

total_camera_days_hourly <- sum(camera_days_vec_hourly)

# Add detection rate per camera day
species_final_hourly <- species_final_hourly %>%
  mutate(
    rate_per_cam_day_hibernacula = count_observed_at_artificial_refuge / camera_days_vec_hourly["Artificial Refuge"],
    rate_per_cam_day_boulder = count_observed_at_boulder / camera_days_vec_hourly["Boulder"],
    rate_per_cam_day_log = count_observed_at_log / camera_days_vec_hourly["Log"]
  )

species_final_hourly <- species_final_hourly %>%
  left_join(
    clean_pt_2_seq_dep_lumped %>%
      dplyr::select(common_name, genus_species, class_grouped) %>%
      distinct(),   # in case there are duplicates
    by = "common_name"
  )

# Then manually patch the missing ones
species_final_hourly <- species_final_hourly %>%
  mutate(
    genus_species = case_when(
      common_name == "Western Meadowlark"    ~ "Sturnella neglecta",
      common_name == "American Pipit"        ~ "Anthus rubescens",
      common_name == "Wren"                  ~ "Troglodytes spp.",
      common_name == "Grey Fox"              ~ "Urocyon cinereoargenteus",
      common_name == "Western Spotted Skunk"         ~ "Spilogale leucoparia",
      common_name == "California Towhee"     ~ "Melozone crissalis",
      common_name == "Common Garter Snake"   ~ "Thamnophis sirtalis",
      common_name == "Cooper's Hawk"         ~ "Astur cooperii",
      TRUE ~ genus_species   # keep existing values
    ),
    class_grouped = case_when(
      common_name %in% c("Western Meadowlark", "American Pipit", "Wren", 
                         "California Towhee", "Cooper's Hawk") ~ "Birds",
      common_name %in% c("Grey Fox", "Spotted Skunk") ~ "Mammals",
      common_name == "Common Garter Snake" ~ "Reptiles",
      TRUE ~ class_grouped   # keep existing values
    )
  )

# Format final table
t1_hourly <- species_final_hourly %>%
  dplyr::select(
    'Class' = class_grouped,
    'Common Name' = common_name,
    'Scientific Name' = genus_species,
    'Detections' = count_observed_at_artificial_refuge,
    '% Visited' = percent_hibernacula_visited,
    'Per Cam.-Day' = rate_per_cam_day_hibernacula,
    'Detections ' = count_observed_at_boulder,
    '% Visited ' = percent_boulders_visited,
    'Per Cam.-Day ' = rate_per_cam_day_boulder,
    'Detections  ' = count_observed_at_log,
    '% Visited  ' = percent_logs_visited,
    'Per Cam.-Day  ' = rate_per_cam_day_log
  ) %>%
  mutate(
    'Detections   ' = Detections + `Detections ` + `Detections  `,
    '% Visited   ' = (
      (`% Visited` * total_features_hourly_vec["Artificial Refuge"]) +
      (`% Visited ` * total_features_hourly_vec["Boulder"]) +
      (`% Visited  ` * total_features_hourly_vec["Log"])
    ) / sum(total_features_hourly_vec),
    'Per Cam.-Day   ' = ((Detections + `Detections ` + `Detections  `) / total_camera_days_hourly),
    across(starts_with("%"), ~ paste0(round(.x, 2), "%")),
    across(starts_with("Per Cam.-Day"), ~ round(.x, 3))
  ) %>%
  arrange(desc(`Detections   `), `Common Name`) %>%
  kable(
    caption = "Unique Hours Present and Percent Feature Visited by Feature Type (Hourly Presence)",
    align = c("l", "l", "l", rep("c", 12)),
    digits = 3
  ) %>%
  kable_styling(
    full_width = FALSE,
    position = "left",
    bootstrap_options = c("striped", "hover", "condensed")
  ) %>%
  add_header_above(c(" " = 3, "Artificial Refuges" = 3, "Boulders" = 3, "Logs" = 3, "Total" = 3))

tryCatch({
  save_kable(t1_hourly, here("misc","species_table_hourly_pt_1_and_2_combined.html"))
  webshot(here("misc","species_table_hourly_pt_1_and_2_combined.html"), here("figures", "species_table_hourly_pt_1_and_2_combined.png"), zoom = 2, vwidth = 1800)
}, error = function(e) message("Table save/screenshot skipped: ", e$message))
```
## Species List: Hourly Presence (By Season)

```{r}
#| label: tbl-species-list-hourly_pt_1_and_2_separated
#| tbl-cap: "Species List from Hourly Presence Data (Parts 1 and 2 Separated). Includes species-level wildlife identifications, sorted by class then total detections."

# ---- Totals per feature_type x study_part ----
total_features_hourly <- hourly_presence_pt_1_and_2 %>%
  distinct(feature_type_methodology, placename, study_part) %>%
  count(feature_type_methodology, study_part, name = "total_features_hourly")

# camera effort per feature_type x study_part
effort_by_ft_part <- feature_type_effort_summary_pt_1_and_2 %>%
  group_by(feature_type_methodology, study_part) %>%
  summarise(total_camera_days = sum(total_camera_days, na.rm = TRUE), .groups = "drop")

# ---- Species summaries per feature_type x study_part ----
species_summary_hourly <- hourly_presence_pt_1_and_2 %>%
  group_by(common_name, feature_type_methodology, study_part) %>%
  summarise(
    count_observed   = sum(n, na.rm = TRUE),
    features_visited = n_distinct(placename),
    .groups = "drop"
  )

# ---- Pivot wide (keep study_part as an id col) ----
species_counts_wide_hourly <- species_summary_hourly %>%
  dplyr::select(common_name, study_part, feature_type_methodology, count_observed) %>%
  pivot_wider(
    id_cols   = c(common_name, study_part),
    names_from  = feature_type_methodology,
    values_from = count_observed,
    values_fill = 0,
    names_prefix = "count_observed_at_"
  ) %>%
  clean_names()

species_features_wide_hourly <- species_summary_hourly %>%
  dplyr::select(common_name, study_part, feature_type_methodology, features_visited) %>%
  pivot_wider(
    id_cols   = c(common_name, study_part),
    names_from  = feature_type_methodology,
    values_from = features_visited,
    values_fill = 0,
    names_prefix = "features_visited_at_"
  ) %>%
  clean_names()

# ---- Pivot totals/effort wide so we can divide per-part ----
total_features_wide <- total_features_hourly %>%
  pivot_wider(
    id_cols   = study_part,
    names_from  = feature_type_methodology,
    values_from = total_features_hourly,
    names_prefix = "total_features_"
  ) %>%
  clean_names()

effort_wide <- effort_by_ft_part %>%
  pivot_wider(
    id_cols   = study_part,
    names_from  = feature_type_methodology,
    values_from = total_camera_days,
    values_fill = 0,
    names_prefix = "cam_days_"
  ) %>%
  clean_names()

# ---- Combine all, keyed by common_name + study_part ----
species_combined_hourly <- species_counts_wide_hourly %>%
  full_join(species_features_wide_hourly, by = c("common_name","study_part")) %>%
  left_join(total_features_wide, by = "study_part") %>%
  left_join(effort_wide, by = "study_part")

# Helper: safely divide (avoid Inf/NaN when denom==0 or NA)
safe_div <- function(num, den) ifelse(is.na(den) | den == 0, NA_real_, num / den)

# ---- Compute per-feature % visited and rates (per-part) ----
species_final_hourly <- species_combined_hourly %>%
  mutate(
    # Percent visited (×100)
    percent_hibernacula_visited = 100 * safe_div(
      features_visited_at_artificial_refuge, total_features_artificial_refuge
    ),
    percent_boulders_visited = 100 * safe_div(
      features_visited_at_boulder, total_features_boulder
    ),
    percent_logs_visited = 100 * safe_div(
      features_visited_at_log, total_features_log
    ),

    # Rates per cam-day
    rate_per_cam_day_hibernacula = safe_div(
      count_observed_at_artificial_refuge, cam_days_artificial_refuge
    ),
    rate_per_cam_day_boulder = safe_div(
      count_observed_at_boulder, cam_days_boulder
    ),
    rate_per_cam_day_log = safe_div(
      count_observed_at_log, cam_days_log
    ),

    # Totals across feature types (per part)
    detections_total = rowSums(across(c(count_observed_at_artificial_refuge,
                                        count_observed_at_boulder,
                                        count_observed_at_log)), na.rm = TRUE),
    total_features_all = rowSums(across(c(total_features_artificial_refuge,
                                          total_features_boulder,
                                          total_features_log)), na.rm = TRUE),
    total_cam_days_all = rowSums(across(c(cam_days_artificial_refuge,
                                          cam_days_boulder,
                                          cam_days_log)), na.rm = TRUE),

    percent_visited_all = 100 * safe_div(
      rowSums(across(c(features_visited_at_artificial_refuge,
                       features_visited_at_boulder,
                       features_visited_at_log)), na.rm = TRUE),
      total_features_all
    ),
    rate_per_cam_day_all = safe_div(detections_total, total_cam_days_all)
  )

# ---- Add scientific names to species_final_hourly ----
species_final_hourly <- species_final_hourly %>%
  left_join(
    clean_pt_2_seq_dep_lumped %>%
      dplyr::select(common_name, genus_species, class_grouped) %>%
      distinct(),
    by = "common_name"
  ) %>%
  # Manually patch missing ones (Part 1 species and incomplete taxonomy)
  mutate(
    genus_species = case_when(
      common_name == "Western Meadowlark"    ~ "Sturnella neglecta",
      common_name == "American Pipit"        ~ "Anthus rubescens",
      common_name == "Wren"                  ~ "Troglodytidae",  # Family name
      common_name == "Grey Fox"              ~ "Urocyon cinereoargenteus",
      common_name == "Western Spotted Skunk" ~ "Spilogale gracilis",
      common_name == "California Towhee"     ~ "Melozone crissalis",
      common_name == "Common Garter Snake"   ~ "Thamnophis sirtalis",
      common_name == "Cooper's Hawk"         ~ "Accipiter cooperii",
      TRUE ~ genus_species
    ),
    # Update Wren common name to "Wren sp."
    common_name = ifelse(common_name == "Wren", "Wren sp.", common_name)
  )

# ---- Build wide table with Part headers (no Study Part column) ----
# Start from species_final_hourly created above

# 1) Keep just the display metrics
metrics_long <- species_final_hourly %>%
  transmute(
    common_name, study_part,
    # Artificial Refuge (AR)
    detections_hib = count_observed_at_artificial_refuge,
    pct_hib       = percent_hibernacula_visited,
    rate_hib      = rate_per_cam_day_hibernacula,
    # Boulders
    detections_bld = count_observed_at_boulder,
    pct_bld        = percent_boulders_visited,
    rate_bld       = rate_per_cam_day_boulder,
    # Logs
    detections_log = count_observed_at_log,
    pct_log        = percent_logs_visited,
    rate_log       = rate_per_cam_day_log,
    # Totals
    detections_tot = detections_total,
    pct_tot        = percent_visited_all,
    rate_tot       = rate_per_cam_day_all
  ) %>%
  tidyr::pivot_longer(
    cols = -c(common_name, study_part),
    names_to = "metric",
    values_to = "value"
  )

# 2) Go wide by study_part so columns repeat under Part 1 and Part 2
metrics_wide <- metrics_long %>%
  tidyr::pivot_wider(
    id_cols = common_name,
    names_from = c(study_part, metric),
    values_from = value,
    names_glue = "part_{study_part}__{metric}"
  )

# --- figure out which metrics actually exist per part ---
present_feats <- metrics_long %>%
  dplyr::group_by(study_part, metric) %>%
  dplyr::summarise(all_na = all(is.na(value) | value == 0), .groups = "drop") %>%
  dplyr::mutate(feature = sub("^(detections|pct|rate)_(.*)$", "\\2", metric)) %>%
  dplyr::group_by(study_part, feature) %>%
  dplyr::summarise(all_zero = all(all_na), .groups = "drop")

# features to drop (all-zero across all metrics)
drop_feats <- present_feats %>%
  dplyr::filter(all_zero) %>%
  dplyr::mutate(prefix = paste0("__", feature)) %>%
  dplyr::pull(prefix)

# 3) Order columns: Part 1 (Hib, Bld, Log, Total) then Part 2 (same order)
col_order <- c(
  # Part 1
  "part_1__detections_hib","part_1__pct_hib","part_1__rate_hib",
  # Part 2
  "part_2__detections_hib","part_2__pct_hib","part_2__rate_hib",
  "part_2__detections_bld","part_2__pct_bld","part_2__rate_bld",
  "part_2__detections_log","part_2__pct_log","part_2__rate_log",
  "part_2__detections_tot","part_2__pct_tot","part_2__rate_tot"
)

# filter col_order so we remove those
col_order <- col_order[!vapply(drop_feats, function(df) grepl(df, col_order), logical(length(col_order)))]


# some columns may be missing if a feature type is absent in a part; keep only those that exist
col_order <- col_order[col_order %in% names(metrics_wide)]

# for safety, no rename yet
col_keep <- intersect(col_order, names(metrics_wide))

# --- rebuild display_tbl with stable sort key (totals across parts) ---
display_tbl <- metrics_wide %>%
  dplyr::left_join(
    dplyr::transmute(
      metrics_wide,
      common_name,
      total_all = rowSums(dplyr::across(tidyselect::matches("^part_\\d+__detections_tot$")), na.rm = TRUE)
    ),
    by = "common_name"
  ) %>%
  # Join scientific names
  dplyr::left_join(
    species_final_hourly %>%
      dplyr::select(common_name, genus_species) %>%
      dplyr::distinct(),
    by = "common_name"
  ) %>%
  dplyr::select(common_name, genus_species, all_of(col_keep), total_all) %>%
  dplyr::arrange(dplyr::desc(total_all), common_name) %>%
  dplyr::select(-total_all) %>%
  dplyr::mutate(
    dplyr::across(tidyselect::matches("__pct_"),
                  ~ round(.x, 2)),  # Remove paste0(..., "%")
    dplyr::across(tidyselect::matches("__rate_"), ~ round(.x, 3))
  ) %>%
  dplyr::rename(`Common Name` = common_name, `Scientific Name` = genus_species)

# --- build leaf (bottom) labels to pass via kable(col.names=...) ---
metric_label_map <- c(
  detections_hib = "Hrs",  pct_hib = "%",  rate_hib = "Rate",
  detections_bld = "Hrs",  pct_bld = "%",  rate_bld = "Rate",
  detections_log = "Hrs",  pct_log = "%",  rate_log = "Rate",
  detections_tot = "Hrs",  pct_tot = "%",  rate_tot = "Rate"
)

leaf_labels <- c(
  "Common Name",
  "Scientific Name",
  vapply(col_keep, function(x) {
    key <- sub("^part_\\d+__", "", x)                 # e.g., "pct_hib"
    lbl <- unname(metric_label_map[key])
    if (is.na(lbl)) key else lbl
  }, character(1))
)

# --- compute feature spans per part from the actual columns kept ---
count_for <- function(part, suf)
  sum(grepl(paste0("^part_", part, "__.*_", suf, "$"), col_keep))

build_feature_spans <- function(part) {
  out <- c()
  n <- count_for(part, "hib"); if (n > 0) out <- c(out, "Artificial Refuge" = n)
  n <- count_for(part, "bld"); if (n > 0) out <- c(out, "Boulders"               = n)
  n <- count_for(part, "log"); if (n > 0) out <- c(out, "Logs"                    = n)
  n <- count_for(part, "tot"); if (n > 0) out <- c(out, "Total"                   = n)
  out
}

feature_header <- c(" " = 2, build_feature_spans(1), build_feature_spans(2))  # 2 for Common Name + Scientific Name

n_part1 <- sum(grepl("^part_1__", col_keep))
n_part2 <- sum(grepl("^part_2__", col_keep))
top_header <- c(" " = 2, if (n_part1 > 0) c("Winter" = n_part1), if (n_part2 > 0) c("Spring" = n_part2))  # 2 for Common Name + Scientific Name

# --- render: leaf labels via col.names; only TWO header rows above ---
t1_hourly <- display_tbl %>%
  kable(
    caption = "Species List from Hourly Presence Data (Parts 1 and 2 Separated). Scientific names follow Wildlife Insights taxonomy (https://www.wildlifeinsights.org/get-started/taxonomy). Hrs = Hours detected; % = Percent of features visited (percentage values); Rate = Detections per camera-day.",
    col.names = leaf_labels,
    align = c("l", "l", rep("c", ncol(display_tbl) - 2)),  # Left-align first two cols (common + scientific), center rest
    digits = 3,
    escape = FALSE
  ) %>%
  kable_styling(full_width = FALSE, position = "left",
                bootstrap_options = c("striped", "hover", "condensed"),
                font_size = 10) %>%
  add_header_above(feature_header) %>%
  add_header_above(top_header) |>
  column_spec((2 + n_part1), border_right = TRUE)  # Adjust border: now 2 columns before Part 2

tryCatch({
  save_kable(t1_hourly, here("misc","species_table_hourly_pt_1_and_2_separated.html"))
  webshot(here("misc","species_table_hourly_pt_1_and_2_separated.html"),
          here("figures","species_table_hourly_pt_1_and_2_separated2.png"),
          zoom = 2, vwidth = 1800)
}, error = function(e) message("Table save/screenshot skipped: ", e$message))

readr::write_csv(display_tbl, here("misc","species_table_hourly_pt_1_and_2_separated.csv"))

```

# Final Figures

Publication-ready versions of manuscript figures and tables, incorporating reviewer feedback.

## Figure 2. Camera Station Map

```{r fig-final-2-map}
#| echo: false
#| fig-cap: "**Figure 2. Camera station map.** Spatial distribution of camera trap stations during (A) the winter study and (B) the spring study. Point shape indicates feature type: square = artificial refuge, triangle = boulder, circle = log. Filled squares indicate artificial refuges sampled in both study periods; open squares were sampled in only one. Landscape shade indicates habitat type. Inset shows the study's location within California. Satellite imagery does not depict the site at the time of study."
#| fig-width: 10
#| fig-height: 10.7

# Reviewer edits applied:
#  - panels stacked vertically with (A)/(B) labels
#  - inset moved into bottom-right of panel A
#  - fill vs. no-fill distinguishes re-used artificial refuges
#  - open shapes (no fill colour) for boulders and logs

# Determine which sites were sampled in both study periods
reuse_lookup <- placenames_for_deployment_map_step1 %>%
  mutate(latitude = round(latitude, 4), longitude = round(longitude, 4)) %>%
  group_by(placename) %>%
  summarise(n_parts = n_distinct(study_part), .groups = "drop") %>%
  mutate(reuse_status = if_else(n_parts > 1, "both", "one"))

# Build a single shape key combining feature type and re-use status
sites_for_final_map <- placenames_for_deployment_map_step1 %>%
  left_join(reuse_lookup, by = "placename") %>%
  mutate(
    # Labels kept short so the whole legend fits on one row; the figure caption
    # spells out what "both" and "one" refer to.
    shape_key = case_when(
      feature_type_methodology == "Artificial Refuge" & reuse_status == "both" ~ "Artificial refuge (both)",
      feature_type_methodology == "Artificial Refuge"                          ~ "Artificial refuge (one)",
      feature_type_methodology == "Boulder"                                    ~ "Boulder",
      TRUE                                                                     ~ "Log"
    ),
    shape_key = factor(shape_key, levels = c(
      "Artificial refuge (both)",
      "Artificial refuge (one)",
      "Boulder",
      "Log"
    ))
  )

# 15 = filled square, 0 = open square, 2 = open triangle, 1 = open circle
final_map_shapes <- c(
  "Artificial refuge (both)" = 15,
  "Artificial refuge (one)"  = 0,
  "Boulder"                  = 2,
  "Log"                      = 1
)

sites_winter_final <- sites_for_final_map %>%
  filter(study_part == 1) %>%
  st_as_sf(coords = c("longitude", "latitude"), crs = 4326)

sites_spring_final <- sites_for_final_map %>%
  filter(study_part == 2) %>%
  st_as_sf(coords = c("longitude", "latitude"), crs = 4326)

# Plotted map extent, used to anchor the inset and the place labels in map
# coordinates rather than in figure fractions.
map_xr <- range(map_rgb_df$x)
map_yr <- range(map_rgb_df$y)

# Adjacent open spaces called out in the manuscript. Esri.WorldImagery carries no
# place labels, so these are drawn explicitly as edge banners: Ellwood Mesa runs
# up the western edge, Coal Oil Point Reserve along the southern edge.
place_labels <- data.frame(
  label = c("Ellwood Mesa Open Space", "Coal Oil Point Reserve"),
  fx    = c(0.018, 0.53),
  fy    = c(0.42,  0.030),
  angle = c(90, 0),
  stringsAsFactors = FALSE
) %>%
  mutate(
    x = map_xr[1] + fx * diff(map_xr),
    y = map_yr[1] + fy * diff(map_yr)
  )

build_final_panel <- function(site_data, panel_title) {
  ggplot() +
    geom_raster(data = map_rgb_df, aes(x = x, y = y, fill = hex)) +
    scale_fill_identity() +
    ggnewscale::new_scale_fill() +
    geom_sf(data = ncos_habitats_filtered, aes(fill = habitat_simple), color = NA, alpha = 0.5) +
    scale_fill_manual(values = habitat_colors, name = "Habitat Type") +
    geom_sf(data = open_space_boundary, aes(color = "NCOS Boundary"),
            fill = NA, linewidth = 1.5) +
    scale_color_manual(values = c("NCOS Boundary" = "yellow"), name = NULL) +
    geom_sf(data = site_data, aes(shape = shape_key), color = "black", size = 3.8, stroke = 1.3) +
    scale_shape_manual(values = final_map_shapes, name = "Feature Type", drop = FALSE) +
    geom_label(data = place_labels,
               aes(x = x, y = y, label = label, angle = angle),
               inherit.aes = FALSE, color = "grey10",
               fill = alpha("grey82", 0.88), label.size = 0,
               size = 3.0, label.padding = unit(0.18, "lines")) +
    # expand = FALSE keeps the panel flush with the raster extent; with the
    # default expansion the axes overshoot the imagery and edge-anchored
    # annotations (the scale bar) land outside the visible map.
    coord_sf(expand = FALSE) +
    labs(title = panel_title) +
    annotation_scale(location = "bl", pad_x = unit(0.22, "in"), pad_y = unit(0.18, "in"),
                     width_hint = 0.2, text_col = "white") +
    theme_minimal(base_size = 11) +
    theme(
      panel.grid = element_blank(),
      axis.title = element_blank(),
      axis.text = element_blank(),
      axis.ticks = element_blank(),
      plot.title = element_text(size = 15, face = "bold", hjust = 0),
      legend.position = "none"
    )
}

panel_a <- build_final_panel(sites_winter_final, "(A) Winter (Feb–Mar 2021)")
panel_b <- build_final_panel(sites_spring_final, "(B) Spring (Apr–May 2021)")

# CA inset
inset_ca <- ggplot() +
  geom_sf(data = map_CA, fill = "gray90", color = "gray60", size = 0.3) +
  geom_point(data = map_centroid_coords, aes(x = longitude, y = latitude),
             color = "red", size = 2.5, shape = 15) +
  theme_void() +
  theme(
    panel.border = element_rect(color = "black", fill = NA, size = 0.5),
    plot.background = element_rect(fill = "white", color = NA)
  )

# Anchor the inset in the map's own coordinate space rather than in figure
# fractions: annotation_custom placement is relative to the plotted map extent,
# so the inset stays flush in the bottom-right corner of panel A regardless of
# how the surrounding panel/legend heights are laid out.
map_xr <- range(map_rgb_df$x)
map_yr <- range(map_rgb_df$y)
pad_x  <- 0.004 * diff(map_xr)
pad_y  <- 0.008 * diff(map_yr)
inset_w <- 0.20 * diff(map_xr)
inset_h <- 0.42 * diff(map_yr)

panel_a_with_inset <- panel_a +
  annotation_custom(
    ggplotGrob(inset_ca),
    xmin = map_xr[2] - pad_x - inset_w, xmax = map_xr[2] - pad_x,
    ymin = map_yr[1] + pad_y,           ymax = map_yr[1] + pad_y + inset_h
  )

# Shared legend built from the full (spring) site set
final_legend_plot <- ggplot() +
  geom_raster(data = map_rgb_df[1, ], aes(x = x, y = y, fill = hex)) +
  scale_fill_identity() +
  ggnewscale::new_scale_fill() +
  geom_sf(data = ncos_habitats_filtered, aes(fill = habitat_simple), color = NA, alpha = 0.5) +
  scale_fill_manual(values = habitat_colors, name = "Habitat Type") +
  geom_sf(data = open_space_boundary, aes(color = "NCOS Boundary"),
          fill = NA, linewidth = 1.5) +
  scale_color_manual(values = c("NCOS Boundary" = "yellow"), name = NULL) +
  geom_sf(data = sites_spring_final, aes(shape = shape_key), color = "black", size = 3.8, stroke = 1.3) +
  scale_shape_manual(values = final_map_shapes, name = "Feature Type", drop = FALSE) +
  # All three legend blocks sit side by side on a single row. Text is sized down
  # so the full row (8 keys + 2 titles) fits inside the figure width without
  # clipping at either end.
  theme(legend.position = "bottom", legend.box = "horizontal",
        legend.title = element_text(size = 9.5, face = "bold"),
        legend.text = element_text(size = 8.5),
        legend.spacing.x = unit(0.12, "cm"),
        legend.key.width = unit(0.42, "cm"),
        legend.margin = margin(2, 2, 2, 2)) +
  guides(
    fill  = guide_legend(order = 1, nrow = 1, title.position = "left", title.vjust = 0.5,
                         override.aes = list(alpha = 0.7, shape = 22, color = NA, size = 3.6)),
    shape = guide_legend(order = 2, nrow = 1, title.position = "left", title.vjust = 0.5,
                         override.aes = list(size = 3.0, stroke = 1.0)),
    color = guide_legend(order = 3, nrow = 1,
                         override.aes = list(fill = NA, linewidth = 1.2))
  )

final_shared_legend <- cowplot::get_legend(final_legend_plot)

# The map has a 2:1 aspect, so each panel is sized to roughly title + map height;
# any extra allotted height would render as dead space beneath the map. The
# single-row legend needs only a thin strip.
fig_final_2 <- plot_grid(
  plot_grid(panel_a_with_inset, panel_b, ncol = 1, rel_heights = c(1, 1)),
  final_shared_legend,
  ncol = 1, rel_heights = c(1, 0.06)
)

fig_final_2

ggsave(here("figures", "FINAL_fig2_camera_station_map.png"), plot = fig_final_2,
       width = 10, height = 10.7, dpi = 300, bg = "white")
```

## Figure 4. Model-Averaged Effect Sizes

```{r fig-final-4-coefplot}
#| echo: false
#| fig-cap: "**Figure 4. Model-averaged effect sizes.** Incidence rate ratios (points) with 95% confidence intervals (bars) for each predictor, averaged across the models within ΔAICc < 2 and weighted by Akaike weights. The dashed line at IRR = 1 marks no difference from the reference condition (artificial refuge, not trail-adjacent, grassland habitat); intervals spanning it are consistent with no effect. Values below 1 indicate fewer detections than the reference, values above 1 more. The intercept is omitted because it is a baseline rate rather than a ratio. Note the log-scaled x-axis."
#| fig-width: 8
#| fig-height: 5

if (exists("avg_display")) {

  coef_plot_data <- avg_display %>%
    # The intercept is a baseline rate, not a ratio, so it has no place on an
    # IRR axis; drop it and keep only the contrast terms.
    filter(!grepl("^Intercept", Predictor)) %>%
    mutate(
      group = case_when(
        Predictor %in% c("Boulder", "Log")     ~ "Feature type\n(vs. artificial refuge)",
        Predictor == "Trail-adjacent"          ~ "Trail adjacency\n(vs. interior)",
        Predictor %in% c("Marsh", "Scrub") ~ "Habitat\n(vs. grassland)",
        TRUE                                   ~ "Other"
      ),
      # Order predictors so related terms sit together on the axis
      Predictor = factor(Predictor,
                         levels = rev(c("Boulder", "Log", "Trail-adjacent",
                                        "Marsh", "Scrub"))),
      # Flag whether the interval excludes 1 (i.e. is distinguishable from null)
      excludes_one = (CI_Lower > 1) | (CI_Upper < 1)
    )

  fig_final_4 <- ggplot(coef_plot_data,
                        aes(x = IRR, y = Predictor)) +
    geom_vline(xintercept = 1, linetype = "dashed", color = "grey40", linewidth = 0.7) +
    geom_errorbarh(aes(xmin = CI_Lower, xmax = CI_Upper),
                   height = 0.18, linewidth = 0.8, color = "grey20") +
    geom_point(aes(fill = excludes_one), shape = 21, size = 4.2,
               color = "black", stroke = 1.0) +
    scale_fill_manual(values = c("TRUE" = "black", "FALSE" = "white"), guide = "none") +
    # IRRs are multiplicative, so a log axis puts "half as many" and "twice as
    # many" at equal visual distance from the null line.
    scale_x_continuous(
      trans  = "log",
      breaks = c(0.125, 0.25, 0.5, 1, 2, 4),
      labels = c("0.125", "0.25", "0.5", "1", "2", "4")
    ) +
    facet_grid(group ~ ., scales = "free_y", space = "free_y", switch = "y") +
    labs(x = "Incidence rate ratio (log scale)", y = NULL) +
    theme_cowplot() +
    theme(
      panel.border     = element_rect(color = "black", fill = NA, linewidth = 0.7),
      strip.background = element_rect(color = "black", fill = "grey88", linewidth = 0.7),
      strip.text.y.left = element_text(angle = 0, size = 10, margin = margin(5, 5, 5, 5)),
      strip.placement  = "outside",
      axis.line        = element_blank(),
      panel.spacing    = unit(0.4, "lines"),
      axis.text        = element_text(size = 12),
      axis.title       = element_text(size = 13)
    )

  ggsave(here("figures", "FINAL_fig4_model_averaged_effects.png"),
         plot = fig_final_4, width = 8, height = 5, dpi = 300, bg = "white")

  # Explicit print(): inside an if() block these are not top-level expressions,
  # so knitr will not auto-print the plot the way it does elsewhere.
  print(fig_final_4)
}
```

## Figure 5. Detection Rates by Feature Type

```{r fig-final-5-detection-rates}
#| echo: false
#| fig-cap: "**Figure 5. Detection rates by feature type.** Average number of wildlife detections per camera trap day across the three feature types in the spring study. Open points denote trail-adjacent (edge) features; filled points denote interior features. Black diamonds indicate mean values and vertical lines show 95% confidence intervals."
#| fig-width: 8
#| fig-height: 6

# Reviewer edits applied:
#  - panel title removed
#  - edge (trail-adjacent) features distinguished by open symbols

fig5_data <- placename_summary_pt_2 %>%
  mutate(site_position = factor(site_position,
                                levels = c("Interior", "Edge (trail-adjacent)")))

fig5_summary <- fig5_data %>%
  group_by(feature_type_methodology_recoded) %>%
  summarise(
    mean = mean(detections_per_camera_day, na.rm = TRUE),
    se   = sd(detections_per_camera_day, na.rm = TRUE) / sqrt(n()),
    n    = n(),
    .groups = "drop"
  ) %>%
  mutate(
    lower = mean - qt(0.975, df = n - 1) * se,
    upper = mean + qt(0.975, df = n - 1) * se
  )

fig_final_5 <- ggplot(fig5_data,
                      aes(x = feature_type_methodology_recoded, y = detections_per_camera_day)) +
  geom_jitter(aes(color = feature_type_methodology_recoded, shape = site_position),
              width = 0.22, alpha = 0.85, size = 3.8, stroke = 1.2) +
  geom_errorbar(data = fig5_summary,
                aes(x = feature_type_methodology_recoded, y = mean, ymin = lower, ymax = upper),
                width = 0.15, color = "black", linewidth = 1, inherit.aes = FALSE) +
  geom_point(data = fig5_summary,
             aes(x = feature_type_methodology_recoded, y = mean),
             shape = 23, size = 6, fill = "black", color = "black", stroke = 1.5,
             inherit.aes = FALSE) +
  scale_color_manual(values = c("Artificial Refuge" = "#fc8d62",
                                "Boulder"           = "#66c2a5",
                                "Log"               = "#8da0cb"),
                     guide = "none") +
  scale_shape_manual(values = c("Interior" = 16, "Edge (trail-adjacent)" = 1),
                     name = "Site position") +
  labs(x = "Feature Type", y = "Detections per camera day") +
  theme_cowplot() +
  theme(
    axis.text.x  = element_text(vjust = 1, size = 14),
    axis.text.y  = element_text(size = 12),
    axis.title.x = element_text(size = 14),
    axis.title.y = element_text(size = 14),
    legend.position = "bottom",
    legend.title = element_text(size = 12),
    legend.text  = element_text(size = 11)
  )

fig_final_5

ggsave(here("figures", "FINAL_fig5_detection_rates_by_feature_type.png"),
       plot = fig_final_5, width = 8, height = 6, dpi = 300, bg = "white")
```

## Figure 6. Species Detection Rates by Habitat Type

```{r fig-final-6-species-habitat}
#| echo: false
#| fig-cap: "**Figure 6. Spring study species detection rates per camera day by habitat type.** Species with more than 50 detections in the study period are displayed individually. All panels share a common y-axis scale."
#| fig-width: 7
#| fig-height: 12

# Reviewer edits applied:
#  - panels stacked in a single column so all share x-axis labels
#  - common y-axis scale across panels
#  - facet strip labels no longer italicised
#  - habitat legend dropped (habitats are labelled directly on the x-axis)

fig_final_6 <- ggplot(resident_habitat_data,
                      aes(x = habitat_type, y = detections_per_camera_day, fill = habitat_type)) +
  geom_col(width = 0.7) +
  facet_wrap(~ common_name, ncol = 1) +
  scale_fill_manual(values = habitat_colors) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.1))) +
  labs(x = "Habitat Type", y = "Detections per Camera Day") +
  theme_cowplot() +
  # Box each panel together with its strip so the species label reads as
  # belonging to the plot beneath it rather than floating between panels.
  theme(
    panel.border     = element_rect(color = "black", fill = NA, linewidth = 0.7),
    strip.background = element_rect(color = "black", fill = "grey88", linewidth = 0.7),
    strip.text       = element_text(size = 13, face = "plain",
                                    margin = margin(5, 5, 5, 5)),
    # theme_cowplot draws its own left/bottom axis lines; the border replaces them.
    # Ticks are matched to the border weight so the y = 0 tick reads flush with
    # the panel edge it abuts (a heavier tick against a thin border makes the
    # sub-pixel rasterisation offset look like a misalignment).
    axis.line        = element_blank(),
    axis.ticks       = element_line(color = "black", linewidth = 0.7),
    panel.spacing    = unit(0.9, "lines"),
    axis.text        = element_text(size = 12),
    axis.title       = element_text(size = 14),
    legend.position  = "none"
  )

fig_final_6

ggsave(here("figures", "FINAL_fig6_species_by_habitat.png"),
       plot = fig_final_6, width = 7, height = 12, dpi = 300, bg = "white")
```

## Figure 7. Species Residency Status by Feature Type

```{r fig-final-7-residency}
#| echo: false
#| fig-cap: "**Figure 7. Species residency status by feature type.** Number of hours (per camera day) in which each species was detected at least once, for resident and transient species across feature types in both the winter and spring studies. Points represent individual species. Species with detection rates below 0.25 hours per camera day are classified as transient. The four resident species are coloured individually; all other species are shown in light blue. Black diamonds show mean detection rates with 95% confidence intervals."
#| fig-width: 13
#| fig-height: 7

# Reviewer edits applied:
#  - "refugia" label updated to Artificial Refuge
#  - legend no longer cut off (wraps onto two rows, extra bottom margin)
#  - "Other" renamed to "Other species" to reduce confusion with "transient"

residency_plot_data <- species_hourly_presence_by_feature_pt_1_and_2 %>%
  mutate(species_fill = forcats::fct_recode(species_fill, "Other species" = "Other"))

final_palette <- extended_palette
names(final_palette)[names(final_palette) == "Other"] <- "Other species"

final_residency_plots <- lapply(seq_along(feature_types), function(i) {
  ft <- feature_types[i]

  p <- ggplot(
    residency_plot_data %>% filter(feature_type_methodology == ft),
    aes(x = residency_status, y = detections_per_camera_day)
  ) +
    geom_jitter(aes(fill = species_fill, shape = species_fill),
                color = "black", width = 0.25, alpha = 0.85, size = 5, show.legend = TRUE) +
    geom_errorbar(
      data = summary_stats_species_hourly_presence_by_feature_pt_1_and_2 %>%
        filter(feature_type_methodology == ft),
      aes(x = residency_status, y = mean, ymin = lower, ymax = upper),
      width = 0.5, color = "black", linewidth = 0.7, inherit.aes = FALSE
    ) +
    geom_point(
      data = summary_stats_species_hourly_presence_by_feature_pt_1_and_2 %>%
        filter(feature_type_methodology == ft),
      aes(x = residency_status, y = mean),
      shape = 9, size = 5, fill = "white", color = "black", stroke = 1.2,
      inherit.aes = FALSE
    ) +
    scale_fill_manual(values = final_palette, name = "Species") +
    scale_shape_manual(values = c(21, 22, 23, 24, 25), name = "Species") +
    scale_y_continuous(limits = c(0, max_y)) +
    labs(title = ft, x = NULL, y = NULL) +
    theme_minimal()

  if (i == 1) p <- p + labs(y = "Avg. # of Hours Detected per Camera Day")
  if (i == ceiling(length(feature_types) / 2)) p <- p + labs(x = "Residency Status")

  p
})

fig_final_7 <- wrap_plots(final_residency_plots, guides = "collect") &
  theme_minimal(base_size = 15) &
  theme(
    axis.text.x  = element_text(size = 16, angle = 0, vjust = 0.5),
    axis.text.y  = element_text(size = 14),
    axis.title.x = element_text(size = 16, face = "bold"),
    axis.title.y = element_text(size = 15, face = "bold"),
    plot.title   = element_text(size = 17, face = "bold", hjust = 0.5),
    legend.position = "bottom",
    legend.box      = "horizontal",
    legend.title    = element_text(size = 14),
    legend.text     = element_text(size = 12),
    plot.margin     = margin(5, 5, 15, 5)
  ) &
  guides(fill  = guide_legend(nrow = 2, byrow = TRUE),
         shape = guide_legend(nrow = 2, byrow = TRUE))

fig_final_7

ggsave(here("figures", "FINAL_fig7_species_residency_by_feature_type.png"),
       plot = fig_final_7, width = 13, height = 7, dpi = 300, bg = "white")
```

## Table 1. Competitive Models (ΔAICc < 2)

```{r tbl-final-competitive}
#| echo: false

# Also write a CSV so the table can be pulled straight into Word/Excel as an
# editable table rather than re-typed from the rendered HTML.
top_display %>%
  dplyr::select(-row_type_flag) %>%
  setNames(c("Model / Predictor", "Beta", "IRR", "Pct_Change", "P_Value", "delta_AICc", "w", "RMSE")) %>%
  readr::write_csv(here("misc", "FINAL_table1_competitive_models.csv"), na = "")

top_display %>%
  dplyr::select(-row_type_flag) %>%
  kable(
    col.names = c("Model / Predictor", "β", "IRR", "% Change", "P-Value", "ΔAICc", "w", "RMSE"),
    digits = 3,
    caption = sprintf("Table 1. Negative binomial models within ΔAICc < 2 of the top-ranked model (n = %d camera stations). Models are ranked by ΔAICc; Akaike weights (w) give the relative probability that each model is the best approximating model in the candidate set. Coefficients are shown as β, incidence rate ratios (IRR), and %% change relative to the reference condition. All models include log(camera hours) as an offset. Reference conditions: artificial refuge, not trail-adjacent, grassland habitat. IRR and %% change are blank for intercept rows, where exp(β) is the baseline detection rate per camera hour rather than a ratio.", nrow(placename_summary_pt_2)),
    align = c("l", "r", "r", "r", "r", "r", "r", "r")
  ) %>%
  kable_styling(full_width = FALSE,
                bootstrap_options = c("striped", "hover", "condensed")) %>%
  row_spec(which(top_display$row_type_flag == "header"),
           bold = TRUE, background = "#d4edda")
```

## Table 2. Model-Averaged Coefficients

```{r tbl-final-model-averaged}
#| echo: false

if (exists("avg_display")) {
  # Round on the way out: write_csv would otherwise emit full floating-point
  # precision, unlike the kable() rendering below which honours digits = 3.
  avg_display %>%
    dplyr::select(Predictor, Estimate, `Std. Error`, IRR, Pct_Change, CI_Lower, CI_Upper) %>%
    mutate(across(where(is.numeric), ~ round(.x, 3))) %>%
    setNames(c("Predictor", "Beta", "SE", "IRR", "Pct_Change", "IRR_CI_Lower", "IRR_CI_Upper")) %>%
    readr::write_csv(here("misc", "FINAL_table2_model_averaged.csv"), na = "")

  avg_display %>%
    dplyr::select(Predictor, Estimate, `Std. Error`, IRR, Pct_Change, CI_Lower, CI_Upper) %>%
    kable(
      col.names = c("Predictor", "β", "SE", "IRR", "% Change", "IRR 95% CI Lower", "IRR 95% CI Upper"),
      digits = 3,
      caption = sprintf("Table 2. Model-averaged coefficients across the %d models within ΔAICc < 2, weighted by Akaike weights (conditional averaging, MuMIn). Standard errors are unconditional and incorporate model selection uncertainty. Reference conditions: artificial refuge, not trail-adjacent, grassland habitat.", length(top_models)),
      align = c("l", "r", "r", "r", "r", "r", "r")
    ) %>%
    kable_styling(full_width = FALSE,
                  bootstrap_options = c("striped", "hover", "condensed"))
} else {
  cat("Model averaging did not converge; see the AICc model comparison section.\n")
}
```

# Appendix: Model Diagnostics
```{r}
# ==============================================================================
# NEGATIVE BINOMIAL MODEL DIAGNOSTICS
# ==============================================================================

# Load additional packages for diagnostics
library(performance)  # for model performance metrics
library(DHARMa)       # for residual diagnostics
library(car)          # for outlier tests and VIF

# 1. OVERDISPERSION TEST
# ==============================================================================
# Test if negative binomial is appropriate vs Poisson
cat("=== OVERDISPERSION ASSESSMENT ===\n")
cat("Raw variance-to-mean ratio:", var(placename_summary_pt_2$total_detections) / mean(placename_summary_pt_2$total_detections), "\n")

# Compare Poisson vs Negative Binomial for best model
poisson_model <- glm(total_detections ~ feature_type_methodology_recoded + trail + habitat_type + offset(log(total_camera_hours)), 
                     family = poisson(), data = placename_summary_pt_2)

# Overdispersion test
overdispersion_test <- performance::check_overdispersion(poisson_model)
print(overdispersion_test)

# 2. MODEL SELECTION DIAGNOSTICS
# ==============================================================================
# Create list of all models for comparison
all_models <- list(
  "Feature Type" = nb_offset_1a,
  "Trail" = nb_offset_1b,
  "Habitat Type" = nb_offset_1c,
  "Feature Type + Trail" = nb_offset_2a,
  "Feature Type + Habitat Type" = nb_offset_2b,
  "Trail + Habitat Type" = nb_offset_2c,
  "Full Model" = nb_offset_3a
)

# Comprehensive model comparison
model_comparison <- data.frame(
  Model = names(all_models),
  AIC = sapply(all_models, AIC),
  BIC = sapply(all_models, BIC),
  LogLik = sapply(all_models, logLik),
  Deviance = sapply(all_models, deviance),
  Theta = sapply(all_models, function(x) x$theta),
  df = sapply(all_models, function(x) x$df.residual)
) %>%
  mutate(
    Delta_AIC = AIC - min(AIC),
    Delta_BIC = BIC - min(BIC),
    AIC_weight = exp(-0.5 * Delta_AIC) / sum(exp(-0.5 * Delta_AIC))
  ) %>%
  arrange(AIC)

print("=== MODEL COMPARISON TABLE ===")
print(model_comparison)

# Likelihood ratio tests for nested models
cat("\n=== LIKELIHOOD RATIO TESTS ===\n")
cat("Feature Type vs Full Model:\n")
print(anova(nb_offset_1a, nb_offset_3a))

cat("\nTrail vs Full Model:\n")
print(anova(nb_offset_1b, nb_offset_3a))

cat("\nHabitat vs Full Model:\n")
print(anova(nb_offset_1c, nb_offset_3a))

# 3. RESIDUAL DIAGNOSTICS
# ==============================================================================
# Use the best model (lowest AIC)
best_model <- all_models[[which.min(sapply(all_models, AIC))]]
cat("\n=== RESIDUAL DIAGNOSTICS FOR BEST MODEL ===\n")
cat("Best model:", names(all_models)[which.min(sapply(all_models, AIC))], "\n")

# DHARMa residuals (recommended for GLMMs and count models)
simulationOutput <- simulateResiduals(fittedModel = best_model, plot = FALSE)

# Create diagnostic plots
par(mfrow = c(2, 2))

# QQ plot of residuals
plotQQunif(simulationOutput, main = "QQ Plot of Scaled Residuals")

# Residuals vs fitted
plotResiduals(simulationOutput, main = "Residuals vs Fitted")

# Additional diagnostic plots
plot(best_model, which = 1, main = "Residuals vs Fitted (Base R)")
plot(best_model, which = 2, main = "Normal Q-Q Plot")

par(mfrow = c(1, 1))

# Formal tests
cat("\nFormal diagnostic tests:\n")
cat("Shapiro-Wilk test for normality of residuals: p =", shapiro.test(residuals(best_model))$p.value, "\n")
cat("Kolmogorov-Smirnov test (DHARMa):\n")
print(testResiduals(simulationOutput))

# 4. OUTLIER AND INFLUENCE DIAGNOSTICS
# ==============================================================================
cat("\n=== OUTLIER AND INFLUENCE DIAGNOSTICS ===\n")

# Cook's distance
cooks_d <- cooks.distance(best_model)
cat("Observations with Cook's D > 4/n:\n")
influential_obs <- which(cooks_d > 4/nrow(placename_summary_pt_2))
if(length(influential_obs) > 0) {
  print(placename_summary_pt_2[influential_obs, c("placename", "total_detections", "feature_type_methodology_recoded")])
} else {
  cat("No highly influential observations detected.\n")
}

# Outlier test
outlier_test <- car::outlierTest(best_model)
cat("\nBonferroni outlier test:\n")
print(outlier_test)

# 5. COLLINEARITY DIAGNOSTICS (for models with multiple predictors)
# ==============================================================================
if(length(coef(best_model)) > 2) {  # More than just intercept + one predictor
  cat("\n=== COLLINEARITY DIAGNOSTICS ===\n")
  
  # VIF (only for models with multiple predictors)
  vif_values <- car::vif(best_model)
  cat("Variance Inflation Factors:\n")
  print(vif_values)
  
  if(any(vif_values > 5)) {
    cat("WARNING: VIF > 5 detected, indicating potential collinearity issues.\n")
  }
}

# 6. GOODNESS OF FIT MEASURES
# ==============================================================================
cat("\n=== GOODNESS OF FIT MEASURES ===\n")

# Pseudo R-squared
null_deviance <- best_model$null.deviance
residual_deviance <- best_model$deviance
mcfadden_r2 <- 1 - (residual_deviance / null_deviance)

cat("McFadden's Pseudo R-squared:", round(mcfadden_r2, 3), "\n")
cat("Theta parameter (dispersion):", round(best_model$theta, 3), "\n")
cat("Standard error of theta:", round(best_model$SE.theta, 3), "\n")

# 7. PREDICTION ACCURACY (if desired)
# ==============================================================================
cat("\n=== PREDICTION ACCURACY ===\n")

# Fitted vs observed
fitted_vals <- fitted(best_model)
observed_vals <- placename_summary_pt_2$total_detections

# Correlation between fitted and observed
correlation <- cor(fitted_vals, observed_vals)
cat("Correlation between fitted and observed:", round(correlation, 3), "\n")

# Mean Absolute Error
mae <- mean(abs(fitted_vals - observed_vals))
cat("Mean Absolute Error:", round(mae, 2), "\n")

# Root Mean Squared Error
rmse <- sqrt(mean((fitted_vals - observed_vals)^2))
cat("Root Mean Squared Error:", round(rmse, 2), "\n")

# 8. SAVE DIAGNOSTIC PLOTS
# ==============================================================================
# Create a comprehensive diagnostic plot
png(here("figures", "model_diagnostics_nb_offset.png"), width = 12, height = 8, units = "in", res = 300)
par(mfrow = c(2, 3))

# Plot 1: Residuals vs Fitted
plot(fitted(best_model), residuals(best_model, type = "pearson"), 
     main = "Residuals vs Fitted", xlab = "Fitted Values", ylab = "Pearson Residuals")
abline(h = 0, col = "red", lty = 2)

# Plot 2: QQ plot
qqnorm(residuals(best_model, type = "pearson"), main = "Normal Q-Q Plot")
qqline(residuals(best_model, type = "pearson"), col = "red")

# Plot 3: Cook's distance
plot(cooks_d, type = "h", main = "Cook's Distance", ylab = "Cook's Distance")
abline(h = 4/nrow(placename_summary_pt_2), col = "red", lty = 2)

# Plot 4: Fitted vs Observed
plot(observed_vals, fitted_vals, main = "Fitted vs Observed",
     xlab = "Observed", ylab = "Fitted")
abline(0, 1, col = "red", lty = 2)

# Plot 5: Histogram of residuals
hist(residuals(best_model, type = "pearson"), main = "Histogram of Residuals",
     xlab = "Pearson Residuals", breaks = 10)

# Plot 6: Scale-location plot
sqrt_abs_resid <- sqrt(abs(residuals(best_model, type = "pearson")))
plot(fitted(best_model), sqrt_abs_resid, 
     main = "Scale-Location Plot", xlab = "Fitted Values", ylab = "√|Residuals|")

dev.off()
par(mfrow = c(1, 1))

cat("\nDiagnostic plots saved to: figures/model_diagnostics_nb_offset.png\n")
```

